---
title: "clinical_epi_diagnosis_I"
output: html_document
---
Between 1992 and 1997, 347 patients with metastatic renal carcinoma were randomised to enter the MRC RE01 trial comparing interferon-alpha (IFA, n=172, variable “group” =1) with medroxyprogesterone acetate (MPA, n=175, variable “group”=0) at 31 centres in the UK. Selected data were available for the following three potential baseline prognostic factors and the primary
outcome: haemoglobin (variable “haem, g dl-1), WHO performance status (variable “who”, range from 0 to 5), time from diagnosis of metastasis to randomization (variable “t_mt”, days), and death (variable 'event', 0=alive and 1=death), respectively. In our exercise, we would like to investigate whether these potential prognostic factors recorded at baseline provide information on the likely outcome of the death in 347 patients.


```{r}


library(conflicted)
library(tidyverse)
library(magrittr)
library(readr)
library(ggplot2)
library(gtsummary)
library(boot)
library(MASS)
library(broom)

conflicts_prefer(
  dplyr::filter(),
  dplyr::select()
)

# 1) import data

# instead of foreign, gpt suggests us to use haven 

install.packages("haven")
library(haven)
df<- read_dta("../data/EXERC_Ewers_mrcre.dta")


```

```{r}
# 2) Describe the structure of dataset and all variables (except variable “ID”) in the dataset? Use e.g. str() , t able () , summary ().

summary(df)

df %>%
  select(group,t_mt, haem, who,event) %>%
  tbl_summary()


```


```{r}
#3) convert days to months (assume there are 30 days in one month).


df <- df %>%
  mutate(t_mt_month = t_mt/30)

```

```{r}
# 4) Check whether “haemoglobin” and “time from diagnosis of metastasis to randomization” follow normal distribution or not. 

qqnorm(df$haem, main="QQ plot: haemoglobin")
qqline(df$haem, col="red")

qqnorm(df$t_mt, main="QQ plot: t_mt")
qqline(df$t_mt, col="red")

shapiro.test(df$haem)
shapiro.test(df$t_mt)

# for haem, p = 0.1584, we cannot reject H0, normally distributed
# for t_mt, p < 2.2e-16, we reject H0, are not normally distributed


```
```{r}
# Since this variable time from diagnosis of metastasis to randomization is not normally distributed, we need to do transformation. 

# First check if there are zeros
summary(df$t_mt)
sum(data$t_mt == 0, na.rm = TRUE)

# We have zeros in the variable. 

df$log_t_mt <- log(df$t_mt + 1)

summary(df$log_t_mt)
# check if we successfully remove 0 )
sum(df$t_mt == 0, na.rm = TRUE)
sum(df$log_t_mt == 0, na.rm = TRUE)


qqnorm(df$log_t_mt)
qqline(df$log_t_mt, col="blue")

shapiro.test(df$log_t_mt)

# now p = 0.007589, transformation is done, variable now is normally distributed. 

```
```{r}

# 5) Calculate the ROC curve and AUC with 95% confidence interval for haemoglobin and interpret the value of AUC. You should install “pROC” package. 
# Use e.g. roc() , plot() , ci.auc()

install.packages("pROC")
library(pROC)

roc_obj <- roc(df$event, df$haem)

plot(roc_obj, col="blue", main="ROC Curve for Haemoglobin")
abline(a=0, b=1, lty=2, col="gray")

auc(roc_obj)
ci.auc(roc_obj)

# H0:AUC=0.5


```
```{r}
# 6)  As plotting ROC curve by using roc() function, you could find the x axis starting from 1. You think it is not good and re-plot ROC curve with x axis equal to 1-specificity and y equal to sensitivity. Use e.g. plot() , abline() 

plot(roc_obj,
     legacy.axes = TRUE,      # transforma x axix Specificity to 1- Specificity
     xlab = "1 - Specificity",
     ylab = "Sensitivity",
     col = "blue",
     main = "ROC Curve (Standard Axes)")

abline(a = 0, b = 1, lty = 2, col = "gray")

```
```{r}
# 7) List operating characteristics of haemoglobin for death, including sensitivity and specificity, and their 95% confidence intervals for each threshold. Use e.g. ci.se() , ci.sp() .

coords(roc_obj, "all", ret=c("threshold", "sensitivity", "specificity"))

ci.se(roc_obj)

ci.sp(roc_obj)

```

```{r}

# 8) Continued to question 7), find the optimal cut-oﬀ point on a basis of Youden’s J statistic.

opt <- coords(roc_obj,
              x = "best",
              best.method = "youden",
              ret = c("threshold", "sensitivity", "specificity", "youden"),
              transpose = FALSE)

opt


plot(roc_obj, legacy.axes=TRUE, main="ROC with Youden optimal cut-off")
abline(0, 1, lty=2, col="gray")

points(opt$specificity, opt$sensitivity, pch=19, col="red")

text(1 - opt$specificity, opt$sensitivity,
     labels = paste0("cut-off=", round(opt$threshold, 2)),
     pos = 4)


```

```{r}

# 9) Continued to question 8), calculate positive predictive value and negative predictive value.
Se <- opt$sensitivity
Sp <- opt$specificity
Prev <- mean(df$event)   # 样本患病率

PPV <- (Se*Prev) / (Se*Prev + (1-Sp)*(1-Prev))
NPV <- (Sp*(1-Prev)) / ((1-Se)*Prev + Sp*(1-Prev))

PPV
NPV


```

```{r}
#10) Continued to question 8), calculate positive likelihood ratio and post-test probability.

```

