---
title: "Data Project NHANES 2025"
author: Yanhang Qiu
output: html_document
---

1.  Familiarize yourself with the NHANES study design and the variables in the dataset by reading the sections "About the NHANES study" and "The data set" of this document. Use the information provided to answer the following questions:

<!-- -->

a)  What are important characteristics of the NHANES study (study design, target population, study objectives, study period, ...)?

**Answer to 1a:**

-   Study design: cross-sectional study - target population: recident non-institutionalized US population (adults and childern)

-   Study objectives: to determine the prevalence of major diseases and risk factors, to design health promotion campaigns, and to establish national reference values (e.g. for height, weight, blood pressure).

-   Study period: This is an ongoing study with data released in bianual cycles.

-   Sampling methods: A four-stage stratified probability cluster sample.

b\. What general categories (such as "demographics", "lab results") can the variables in the dataset be sorted into?

**Answer to 1b:**

The variables in the dataset can be categorized into groups such as demographic data, examination data,  dietary data, lifestyles, disease, family history, selfrated_health, depression, environment and toxic,

```{r 1b, message=FALSE, warning=FALSE}

data <- read.csv2("../data/NHANES1.csv",row.names = 1)
data_raw <- data

library(dplyr)
categories <- list(
  demographic = c("seqn","age","educ","martlst","male","ethnic","increl","age.cat","age_bin"),
  examination  = c("weight","height","bmi","rr_sys","rr_dia","bmi_cat","hdl","logHdl"),
  dietary = c("rdyfood_prvmo","frzfood_prvmo","milk_month"),
  lifestyle =c("alc_lft", "drnkprd_prv12mo", "cigsprd_prv30d", "sleep_dur", "sleep_probl", "hrsworked_prvwk", "wrkt_irreg", "workpollut","smokstat","smoke_bin","cannab_ever","harddrg_ever","jobstat_lwk"),
  disease =c("heartdis_ever","diab_lft", "livdis_ever", "arthrit_ever", "stroke_ever", "cancer_ever", "cbronch_now", "lungpath_ever","asthma_ever","asthma_now","ovrwght_ever","livdis_now","hivpos"),
  fam_history =c("rel_heartdis", "rel_diab", "rel_asthma"),
  selfrated_health = c("srhgnrl", "srphbad_prv30d", "srmhbad_prv30d", "adlimp_prv30d"),
  depression =c ("phq1", "phq2", "phq3", "phq4", "phq5", "phq6", "phq7", "phq8","phq9"),
  envir_and_toxic = c("cd", "pb", "hg")
)

categories$uncategorised_var <- setdiff(names(data), unlist(categories))
cat_df <- stack(categories)
names(cat_df) <- c("variable", "category")

summary_table <- data.frame(
  variable = names(data),                 
  type = sapply(data, class),
  missing = colSums(is.na(data)),
  stringsAsFactors = FALSE
)

summary_table <- merge(summary_table, cat_df, by = "variable", all.x = TRUE)


summary_table %>%
  filter(category == "uncategorised_var") %>%
  select(variable, variable)

summary_table
```



2\. Using the subsample you have chosen, describe the US population with regards to:

a\. Demographic characteristics. How can the strange age distribution be explained? To deal with this problem, recode the age variable into the following categories:18-34, 35-49, 50-64, 65-79, 80 or higher.

```{r 2a, message=FALSE, warning=FALSE}
library(ggplot2) 

plot_age <- ggplot(data, aes(x = age)) +
  geom_histogram(fill = "pink", color = "white",bins = 30) +
  labs(title = "age",x = "Age",y = "Count") +
  theme_minimal()
```

```{r 2a, message=FALSE, warning=FALSE}
library(dplyr)
data <- data %>%
  mutate(
    age_group = case_when(
      age < 18 ~ "0-17",
      age >= 18 & age < 35 ~ "18-34",
      age >= 35 & age < 50 ~ "35-49",
      age >= 50 & age < 65 ~ "50-64",
      age >= 65 & age < 80 ~ "65-79",
      age >= 80 ~ "80+"
    )
  )
```

```{r 2a, message=FALSE, warning=FALSE}
plot_age_cate <-ggplot(data, aes(x=factor(age_group))) + geom_bar(fill = "pink") + 
  labs(title = "age_group", x = "Age Group", y = "Count") +
  theme_minimal()


library(patchwork)
(plot_age | plot_age_cate) 
```

**Answer to 2a:**

The age distribution in the sample shows limited representativeness of the U.S. population, displaying an approximately uniform count across all age intervals. Despite the complex four-stage sampling design, potential biases may still arise — for example, due to substantial differences in age composition across districts, or because older participants tend to be less willing to respond compared to younger individuals. By grouping participants into age categories, we can reweight each group so that the overall age distribution in the sample better reflects that of the U.S. population.



b\. When asked about Marital status, some of the participants refused to answer while some didn't know which category they belonged. Hence they were coded differently. Take that into account and recode the variable martlst.

```{r 2b, message=FALSE, warning=FALSE}

data <- data %>%
  mutate(
    martlst_clean = na_if(martlst, 77),
    martlst_clean = na_if(martlst_clean, 99)
  )

table(data$martlst)
table(data$martlst_clean, useNA = "ifany")

```

**Answer to 2b:**

Before recoding martlst the requency of 77 and 99 are 267 and 1 respectively. After recoding the missing value into NA, there are 268 NAs in total.





c\. Self-rated health. Looking at the results of your descriptive analysis, what do you have to consider when interpreting the results?

```{r 2c1, message=FALSE, warning=FALSE}

library(ggplot2)
library(scales)

p1 <- ggplot(data, aes(x = factor(srhgnrl), y = after_stat(prop), group = 1)) +
  geom_bar(aes(y = after_stat(prop)), fill = "lightblue") +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "srhgnrl")

p1

```

**Answer to 2c:**

According to the bar chart, most participants reported their general health as “good,” “very good,” or “excellent,” suggesting an overall positive perception of health in this sample. However, a considerable proportion of responses for the self-rated health item was missing (over 10%). Such missingness may be systematic rather than random, potentially resulting from factors such as privacy concerns, social desirability bias, or poor physical and mental health conditions that prevent individuals from completing self-rating questions.

To examine this potential bias, we hypothesized that the missingness pattern differs across age, gender, income and used chi-square analyses to test whether missingness in the self-rated health was systematically associated with these characteristics.

```{r 2c2, message=FALSE, warning=FALSE}

library(dplyr)
library(broom)

self_health <- data %>%
  select(age_group, male, increl, srhgnrl) %>%
  mutate(
    missing_self = as.integer(is.na(srhgnrl))
  )

by_gender <- self_health %>%
  group_by(male) %>%
  summarise(n = n(),
            n_missing = sum(missing_self),
            pct_missing = round(100 * n_missing / n, 1),
            .groups = "drop")

by_age <- self_health %>%
  group_by(age_group) %>%
  summarise(n = n(),
            n_missing = sum(missing_self),
            pct_missing = round(100 * n_missing / n, 1),
            .groups = "drop")

by_increl <- self_health %>%
  group_by(increl) %>%
  summarise(n = n(),
            n_missing = sum(missing_self),
            pct_missing = round(100 * n_missing / n, 1),
            .groups = "drop")

by_gender; by_age; by_increl


chisq_gender <- chisq.test(table(self_health$male,       self_health$missing_self))
chisq_age    <- chisq.test(table(self_health$age_group,  self_health$missing_self))
chisq_increl <- self_health %>%
  filter(!is.na(increl)) %>%
  with(chisq.test(table(increl, missing_self)))

chisq_gender
chisq_age
chisq_increl


```

**Answer to 2c (continue) :**

Chi-square tests were conducted to examine whether missingness in the self-rated health variable was randomly distributed across demographic and socioeconomic characteristics. The results indicate that missing self-rated health information was significantly associated with sex (χ² = 13.52, p \< 0.001), age group (χ² = 40.49, p \< 0.001), and family income relative to the poverty line (χ² = 23.96, p \< 0.001).

These findings suggest that missingness in the self-rated health variable is systematic rather than random. In particular, participants who are male, older, or have lower family income were more likely to have missing responses. As a result, complete-case analyses may disproportionately exclude individuals with potentially poorer health and socioeconomic disadvantage, leading to an overestimation of overall health status in the analytic sample. Therefore, the missing data mechanism is unlikely to be Missing Completely at Random (MCAR), and appropriate methods such as multiple imputation or inverse probability weighting should be considered to reduce selection bias.



3.Diabetes and ethnicity

a\. How is the diabetes status distributed in your data set? Summarize the 2 diabetes groups in a single group and recode the variable for diabetes status into two groups: non-diabetes and diabetes. Give an interval estimate for diabetes in your data set.


```{r 3a, message=FALSE, warning=FALSE}

library(dplyr)
library(broom)
library(binom)

table(data$diab_lft, useNA = "ifany")
barplot(table(data$diab_lft),
        main = "Original Diabetes Variable (diab_lft)",
        ylab = "Count", col = "grey80")


data <- data %>%
  mutate(
    diabetes_status = case_when(
      diab_lft == 1 ~ "non_diabetes",
      diab_lft %in% c(2, 3) ~ "diabetes",
      TRUE ~ NA_character_
    )
  )

table(data$diabetes_status, useNA = "ifany")


summary_diabetes <- data %>%
  filter(!is.na(diabetes_status)) %>% 
  count(diabetes_status) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1)
  )

summary_diabetes


counts <- table(data$diabetes_status)
perc <- round(counts / sum(counts) * 100, 1)

bp <- barplot(
  counts,
  col = "grey80",
  ylim = c(0, max(counts) * 1.15),
  main = "Distribution of Diabetes Status",
  ylab = "Count"
)

text(
  x = bp,
  y = counts,
  labels = paste0(counts, "\n(", perc, "%)"),
  pos = 3, cex = 1.0
)


x <- sum(data$diabetes_status == "diabetes", na.rm = TRUE)
n <- sum(!is.na(data$diabetes_status))

# Clopper–Pearso
CI_CP <- binom.test(x, n, conf.level = 0.95)
tidy(CI_CP)

# Wilson CI
CI_Wilson <- binom.confint(x, n, conf.level = 0.95, methods = "wilson")
CI_Wilson




```

**Answer to 3a :**

After recoding diabetes status into two categories (“diabetes” and “non-diabetes”), a total of 4,692 participants had valid information. Among them, 602 individuals (12.8%) were classified as having diabetes, while 4,090 individuals (87.2%) were classified as non-diabetic, indicating that diabetes is relatively uncommon in this study population.

Based on our data, the estimated proportion of individuals with diabetes is about 12.8%. The CI are consistent using two methods : Wilson and Clopper-Pearson methods. The 95% confidence interval ranges approximately from 11.9% to 13.8 in both methods, meaning that if you repeated this sampling many times, 95% of such intervals would contain the true population proportion of diabetes.

However, diabetes status was missing for 308 individuals. Since the analysis excluded these cases, the estimated prevalence and confidence interval reflect only the observed sample and may not fully represent the entire study population.



b\. Work with the recoded variable. Use an appropriate statistical test to test the  relation between diabetes status and ethnicity. Interpret the results of the test.

```{r}

data$diabetes_status <- factor(data$diabetes_status)
data$ethnic <- factor(data$ethnic)

tab <- table(data$diabetes_status, data$ethnic)
tab

chisq.test(tab)

```

```{r}
library(dplyr)

data <- data %>%
  mutate(
    diabetes_missing = ifelse(is.na(diabetes_status), "Missing", "Observed"),
    ethnic = factor(ethnic)
  )


tab_miss <- table(data$diabetes_missing, data$ethnic)
tab_miss

chisq.test(tab_miss)

```

**Answer to 3b :**

As diabetes_status is binomial variable and ethnicity is a nominal variable, we use chi-square test to exam their relationship. Results show that there is significant association (p = 7.296e-08 \<0.05) between diabetes status and ethinicty.

The chi-square test examining the association between diabetes status missingness and ethnicity yielded a non-significant result (χ²(3) = 1.05, p = 0.79). Therefore, the missing data on diabetes status are likely to be missing at random with respect to ethnicity, and excluding these missing cases from the analysis is unlikely to bias the estimated association between diabetes status and ethnicity.


4\. Weekly working hours and self-evaluated health status

a\. The variable `hrsworked_prvwk` codes the number of weekly working hours. 77777 or 99999 are missing values. Recode these missing values with a proper value, e.g. NA.

```{r message=FALSE, warning=FALSE}

data <- data %>%
  mutate(
    hrsworked_prvwk_clean = na_if(hrsworked_prvwk, 77777),
    hrsworked_prvwk_clean = na_if(hrsworked_prvwk_clean, 99999)
  )

sum(is.na(data$hrsworked_prvwk_clean))        
mean(is.na(data$hrsworked_prvwk_clean)) * 100 


data$missing_hrs <- ifelse(is.na(data$hrsworked_prvwk_clean), "Missing", "Observed")
data$missing_srh <- ifelse(is.na(data$srhgnrl), "Missing", "Observed")

tab <- table(data$missing_hrs, data$missing_srh)
tab
chisq.test(tab)

```

**Answer to 4a :**

In variable `hrsworked_prvwk` there are 2 77777, 1 99999 and 2474 NA values. After recoding 77777 and 99999 to NA, there are 2477 missing values in total.

Although the missingness of weekly working hours and self-rated health appeared to be random (χ² = 0.71, p = 0.399), the proportion of missing data for weekly working hours was very high (49.54%).

To minimize potential information loss and improve the robustness of the analyses, multiple imputation was performed. After multiple imputation, the newly created dataset contained no missing values (number of missing values = 0). In the following analyses, both the original (non-imputed) and imputed datasets were retained and analyzed in parallel for comparison.

```{r}
library(mice)


vars <- c("srhgnrl", "hrsworked_prvwk_clean", "age", "male", "bmi")
imp_data <- data[, vars]

pred <- make.predictorMatrix(imp_data)

pred["hrsworked_prvwk_clean", "srhgnrl"] <- 0  
pred["srhgnrl", "hrsworked_prvwk_clean"] <- 0  

pred[c("srhgnrl", "hrsworked_prvwk_clean"), c("srhgnrl", "hrsworked_prvwk_clean")]

imp <- mice(imp_data, m = 5, method = "pmm", predictorMatrix = pred, seed = 123)

completed1 <- complete(imp, 1)
head(completed1)

colSums(is.na(completed1))

imp_long <- complete(imp, "long", include = FALSE)
head(imp_long)

```


b\. Recode the variable `hrsworked_prvwk` into categories [,40), [40,) weekly working hours. Number or name the categories properly. (Hint: Set the new variable as a factor variable, if necessary)

```{r 4b message=FALSE, warning=FALSE}

library(dplyr)

data <- data %>%
  mutate(
    hrsworked_prvwk_cat = case_when(
      is.na(hrsworked_prvwk_clean) ~ NA_character_,   
      hrsworked_prvwk_clean < 40 ~ "< 40",
      hrsworked_prvwk_clean >= 40 ~ ">= 40"
    )
  )


table(data$hrsworked_prvwk_cat,useNA = "ifany")


```
```{r}

completed1 <- complete(imp, 1)

completed1 <- completed1 %>%
  mutate(
    hrs_cat = cut(hrsworked_prvwk_clean,
                  breaks = c(0, 40, Inf),
                  labels = c("<40", "≥40"),
                  right = FALSE)
  )

table(completed1$hrs_cat,useNA = "ifany")
```

**Answer to 4b :**

Before multiple imputation, there are 846 people working less than 40 hours previous week, 1677 people working more than 40 hours previous week, and still, there are 2477 missing values.

After multiple imputation and subsequent categorization, there were 1,743 individuals working less than 40 hours in the previous week and 3,257 individuals working 40 hours or more. No missing values remained in the imputed dataset.


c\. Plot weekly working hours against self-evaluated health status using mosaic plot. Looking at the plot, what would you tell about the relation between these two variables.


```{r 4c message=FALSE, warning=FALSE }

mosaicplot(table(data$srhgnrl, data$hrsworked_prvwk_cat), color = TRUE, main = "Mosaic plot: self-rated health vs weekly working hours (with missing values)") 


mosaicplot(table(completed1$srhgnrl, completed1$hrs_cat),
           color = TRUE,
           main = "Mosaic plot: self-rated health vs weekly working hours (imputed data)")

```
**Answer to 4c :** According to the mosaic plot, the distribution of self-rated health levels is similar across the two working-hour groups (\< 40 vs. ≥ 40 hours). Visually, there is no substantial difference between the plots before and after multiple imputation.



d\. Use an appropriate statistical test to test for the statistical significance of this relation. Interpret the results of the test.

```{r 4d message=FALSE, warning=FALSE }

d1 <- subset(data, !is.na(srhgnrl) & !is.na(hrsworked_prvwk_cat))
tab1 <- table(d1$srhgnrl, d1$hrsworked_prvwk_cat)
chisq.test(tab1)

tab2 <- table(completed1$srhgnrl, completed1$hrs_cat)
chisq.test(tab2)



d3 <- subset(data, !is.na(srhgnrl) & !is.na(hrsworked_prvwk_clean))
cor.test(d2$hrsworked_prvwk_clean, d2$srhgnrl,
         method = "spearman", exact = FALSE)

d4 <-cor.test(completed1$hrsworked_prvwk_clean, 
         as.numeric(as.character(completed1$srhgnrl)), 
         method = "spearman", 
         exact = FALSE)


d3;d4

```


**Answer to 4b :**

Due to concerns about the large amount of missing data, both weekly working hours and self-rated health were imputed.
Before imputation, no significant association was observed (χ² = 3.25, p = 0.52; ρ = –0.038, p = 0.082).

After imputation, the results remained non-significant (χ² = 8.28, p = 0.082; ρ = –0.008, p = 0.59).
Although the significance levels slightly differed before and after imputation, both analyses consistently indicated no significant relationship between the two variables.

Therefore, the influence of missing data on the overall conclusion appears minimal, and the association between weekly working hours and self-rated health remains statistically non-significant.




5\. Blood mercury level (umol/L) and gender

a\. Describe and plot the distribution of blood mercury level in men and women.

```{r 5a message=FALSE, warning=FALSE  }


hg_by_male <- data %>%
  group_by(male) %>%
  summarise(
    mean_hg = mean(hg, na.rm = TRUE),
    sd_hg   = sd(hg, na.rm = TRUE),
    min_hg  = min(hg, na.rm = TRUE),
    max_hg  = max(hg, na.rm = TRUE),
    p25_hg  = quantile(hg, 0.25, na.rm = TRUE),
    p50_hg = median(hg, na.rm = TRUE),
    p75_hg  = quantile(hg, 0.75, na.rm = TRUE),
    n       = sum(!is.na(hg)),
    n_missing = sum(is.na(hg)),
  )

hg_by_male %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

hist(data$hg, col = 2, breaks = 80)
```

**Answer to 5a:**

Through descriptive analysis, we observed that blood mercury levels among participants varied widely, ranging from approximately 0.5 to more than 200 µg/L. The distribution of blood mercury levels was skewed toward the lower range, with 75% of men below 9.8 µg/L and 75% of women below 9.1 µg/L, suggesting that most participants had comparatively low exposure. 

However, a considerable number of individuals were still exposed to elevated blood mercury levels, posing potential health risks that warrant further investigation. In this case, we categorized the blood mercury variable into clinically meaningful risk groups based on internationally recognized guidelines from WHO, CDC, ATSDR, and EPA. This categorization provides a clearer interpretation of exposure severity and allows for more informative and interpretable visualizations.

```{r 5a message=FALSE, warning=FALSE}
library(dplyr)

data <- data %>%
  mutate(
    hg_risk_group = case_when(
      hg < 5                        ~ "<5 µg/L (Background level)",
      hg >= 5  & hg < 15            ~ "5–15 µg/L (Mild elevation)",
      hg >= 15 & hg < 50            ~ "15–50 µg/L (Moderate elevation)",
      hg >= 50 & hg < 100           ~ "50–100 µg/L (High exposure)",
      hg >= 100                     ~ "≥100 µg/L (Potential toxicity)",
      TRUE                          ~ NA_character_
    ),
    hg_risk_group = factor(
      hg_risk_group,
      levels = c(
        "<5 µg/L (Background level)",
        "5–15 µg/L (Mild elevation)",
        "15–50 µg/L (Moderate elevation)",
        "50–100 µg/L (High exposure)",
        "≥100 µg/L (Potential toxicity)"
      ),
      ordered = TRUE
    )
  )

library(dplyr)

n_total <- nrow(data)
n_valid <- sum(!is.na(data$hg_risk_group))

risk_summary <- data %>%
  mutate(group = ifelse(is.na(hg_risk_group), "Missing", as.character(hg_risk_group))) %>%
  count(group, name = "n") %>%
  mutate(
    pct_including_missing = round(n / n_total * 100, 1),   
    pct_excluding_missing = round(ifelse(group == "Missing", NA, n / n_valid * 100), 1) 
  ) %>%

  mutate(group = factor(group,
                        levels = c("<5 µg/L (Background level)",
                                   "5–15 µg/L (Mild elevation)",
                                   "15–50 µg/L (Moderate elevation)",
                                   "50–100 µg/L (High exposure)",
                                   "≥100 µg/L (Potential toxicity)",
                                   "Missing"),
                        ordered = TRUE)) %>%
  arrange(group)

risk_summary


library(ggplot2)
ggplot(
  data %>% filter(!is.na(hg_risk_group)),
  aes(x = hg_risk_group)
) +
  geom_bar(fill = "grey20") +
  labs(
    x = "Blood Mercury Level Category",
    y = "Number of Participants",
    title = "Distribution of Blood Mercury Risk Levels (Missing Excluded)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))



```

**Answer to 5a (continued) :**

The bar chart illustrates the distribution of blood mercury levels after excluding missing observations. The majority of participants (approximately 55.8%) fall within the background exposure range (\<5 µg/L). Notably, 72 individuals (approximately 1.7%) exhibited blood mercury concentrations exceeding 50 µg/L, placing them in the high-exposure or potentially toxic risk category.




b\. Use both parametric and non-parametric statistical tests to test for the statistical significance the relation between blood mercury level and gender. Interpret the results of the tests.

**Answer to 5b :**

Hg is a continuous variable, while male is a binary variable. Normally, when blood mercury levels follow a normal distribution, a parametric test — the independent samples t-test — is appropriate to examine whether blood mercury levels differ between males and females. When the assumption of normality is violated, the non-parametric Mann–Whitney test should be used instead.

The test of normality indicated that Hg was not normally distributed in either group. By comparing the histograms, we observed that both distributions were similarly and heavily right-skewed, which was further supported by the skewness statistics. 

However, given the very large sample size (over 2,000 observations per group), the t-test remains relatively robust to this deviation from normality. The F-test for homogeneity of variances revealed unequal variances between the two groups. Considering all these factors, the Welch’s t-test (which adjusts for unequal variances) or the Mann–Whitney test would be appropriate choices.

```{r}

shapiro.test(data$hg[data$male == 1])
shapiro.test(data$hg[data$male == 0])

hist(data$hg[data$male == 1],
     main = "hg - male",
     xlab = "hg",
     col = "lightblue",
     breaks = 30)

hist(data$hg[data$male == 0],
     main = "hg - female",
     xlab = "hg",
     col = "pink",
     breaks = 30)

library(e1071)
skewness(data$hg[data$male == 1], na.rm = TRUE)
skewness(data$hg[data$male == 0], na.rm = TRUE)


var.test(hg ~ male, data = data)

t.test(hg ~ male, data = data, var.equal = FALSE,conf.level = 0.99)
wilcox.test(hg ~ male, data = data, conf.level = 0.99,conf.int = TRUE) 



sum(is.na(data$hg))

```

**Answer to 5b (continued) :**
Results from both Welch’s t-test (testing mean differences) and the Mann–Whitney U test (testing distributional differences) are consistent in showing no statistically significant difference between the two groups at the 95% confidence level.

Therefore, there is no statistically significant difference in blood mercury levels between men and women.


6.BMI and HDL

a\. How strong is the relationship between BMI and HDL (the "good" cholesterol)? Is it significant? How much of the variation in HDL can be explained (in a statistical sense) by variation in BMI?

```{r}

plot(data$bmi, data$hdl,
     xlab = "BMI", ylab = "HDL",
     main = "Scatterplot of BMI vs HDL")
abline(lm(hdl ~ bmi, data = data), col = "red")


boxplot(data$hdl)
boxplot(data$bmi)

cor(data$hdl, data$bmi,use="complete.obs")
cor.test(data$hdl, data$bmi)

cor(data$hdl, data$bmi, method = "spearman",use="complete.obs")
cor.test(data$hdl, data$bmi, method = "spearman")


model <- lm(hdl ~ bmi, data = data)
summary(model)


```

**Answer to 6a :**

The scatterplot shows that HDL generally decreases as BMI increases, but the points are widely dispersed, indicating that the linear trend is present but weak. The boxplots also reveal several outliers. To obtain more robust and stable results, both Pearson’s r and Spearman’s ρ were therefore examined.

The Pearson correlation between BMI and HDL was –0.28 (95% CI: –0.31 to –0.25), indicating a weak negative linear association. The relationship was highly statistically significant (t = –19.1, df = 4364, p < 2.2×10⁻¹⁶). In addition, the Spearman rank correlation was –0.33 (p < 2.2e-16), suggesting a slightly stronger monotonic decreasing pattern. Together, both measures support a statistically significant but relatively weak negative association between BMI and HDL.

The simple linear regression yielded an R² of 0.077, meaning that approximately 7.7% of the variation in HDL levels can be statistically attributed to variation in BMI. This indicates that BMI alone has limited explanatory power, and most of the variation in HDL must be explained by other factors.



b\. Does the relationship between HDL and BMI change when you adjust for age (categorized)? Interpret the coefficients of the resulting model (when you mean-center BMI before fitting the model, you can also interpret the intercept). Would you say that BMI has a clinically relevant impact on HDL, according to your model?


```{r}

model_controlled <- lm(hdl ~ bmi + age, data = data)
summary(model_controlled)


data$bmi_c <- data$bmi - mean(data$bmi, na.rm = TRUE)
data$age_c <- data$age - mean(data$age, na.rm = TRUE)
model_controlled <- lm(hdl ~ bmi_c + age_c, data = data)
summary(model_controlled)


model_interaction <- lm(hdl ~ bmi_c * age_c, data = data)
summary(model_interaction)
hdl ~ bmi_c + age_c + bmi_c:age_c


```
**Answer to 6b :**

After adding age as a control variable, BMI remained a statistically significant negative predictor of HDL (β = –0.016, p < 2e–16). Its coefficient became slightly more negative, suggesting that controlling for age reduced confounding and revealed a more robust independent association between BMI and HDL. Model fit improved modestly: the R-squared increased from 0.0772 to 0.0862, and the adjusted R-squared showed a similar pattern (from 0.0770 to 0.0857), indicating a genuine though limited improvement in explanatory power.

After mean-centering BMI and age, the intercept became interpretable: the centered model yielded an intercept of 1.355, representing the expected HDL level for an individual with average BMI and average age. This provides a meaningful reference point for understanding the conditional effects of BMI and age.

We also explored whether age modifies the effect of BMI on HDL by including a BMI × age interaction term. The interaction was not statistically significant (β = –8.4×10⁻⁵, p = 0.0746), indicating that the strength of the BMI–HDL association does not meaningfully differ across age levels.

Although BMI showed strong statistical significance, the magnitude of its effect was small. Each one-unit increase in BMI was associated with only a 0.016 mmol/L decrease in HDL, meaning that even a 10-unit difference in BMI would change HDL by only ~0.16 mmol/L—well within the typical biological variability of HDL (often ±0.3–0.5 mmol/L).

Thus, while the association is statistically robust, its clinical relevance appears limited in this dataset.

In addition to evaluating the regression coefficients, it is also important to consider the potential impact of missing data.

```{r}

# 查看缺失数量
sum(is.na(data$hdl))
sum(is.na(data$bmi))
sum(is.na(data$age))
sum(!complete.cases(data[, c("hdl", "bmi", "age")]))


hdl_missing_n <- sum(is.na(data$hdl))
bmi_missing_n <- sum(is.na(data$bmi))
age_missing_n <- sum(is.na(data$age))

hdl_missing_p <- round(mean(is.na(data$hdl)) * 100, 1)
bmi_missing_p <- round(mean(is.na(data$bmi)) * 100, 1)
age_missing_p <- round(mean(is.na(data$age)) * 100, 1)

data.frame(
  Variable = c("hdl", "bmi", "age"),
  Missing_N = c(hdl_missing_n, bmi_missing_n, age_missing_n),
  Missing_Percent = c(hdl_missing_p, bmi_missing_p, age_missing_p)
)

data$missing_hdl <- ifelse(is.na(data$hdl), 1, 0)
data$missing_bmi <- ifelse(is.na(data$bmi), 1, 0)


t.test(age ~ missing_hdl, data = data)
t.test(age ~ missing_bmi, data = data)

```

**Answer to 6b (continued) :**

In the current analysis, 634 observations were excluded due to missing data. Both BMI and HDL had a considerable amount of missingness (574 and 290 cases, corresponding to 11.5% and 5.8%, respectively), whereas age was complete.

Correlation analyses revealed that missingness in both HDL and BMI was significantly associated with age. Specifically, the mean age was 50.0 years in the HDL-missing group versus 47.2 years in the non-missing group (t = -3.18, p = 0.0015), and 52.6 years versus 47.2 years for BMI (t = -4.32, p < 0.001). These findings suggest that the missingness mechanism was not completely at random (approximately MAR) with older participants exhibiting a higher likelihood of missing data. 

This pattern of informative missingness likely attenuated the observed BMI–HDL association. Consequently, the true underlying relationship may be stronger than what is observed here, and a clinically meaningful effect could emerge if missing data were handled appropriately, for example through multiple imputation.


c\. Try to find a better model to predict HDL by including more covariates. Select a number of candidate covariates which in your opinion may be related to HDL, and then choose a model selection strategy and a criterion/test for comparing models. Describe the model with the best fit according to your search, and interpret the model coefficients.

**Answer to 6c:**

The modeling process includes 7 steps:
Step 0. candidate covariates
Step 1. Data Preparation
Step 2. Exploratory Data Analysis (EDA)
Step 3. Specification of the Initial Full Model
Step 4. Model Selection - AIC
Step 5. Assessment of Multicollinearity
Step 6. Model Diagnostics and Assumption Checking
Step 7. Predictive Performance Evaluation and Interpretation


Step 0. candidate covariates
```{r}

library(dplyr)
vars <- c(
  "age",
  "male", 
  "ethnic",
  "educ",
  "increl",
  
  "bmi",
  "rr_sys",
  "hdl",
  "hg",
  "cd",
  
  "sleep_dur",
  "drnkprd_prv12mo",
  "cigsprd_prv30d",
  
  "phq2"   # PHQ-9 item 2: depressed/hopeless
)

```
A set of 13 candidate variables was selected based on previous epidemiological evidence, biological plausibility


Step 1. Data Preparation

```{r}
summary(data[, vars])

data_sub <- data[ , vars]

data_sub <- data_sub %>%
  mutate(
    
    male = factor(male,
                  levels = c(FALSE, TRUE),
                  labels = c("Female", "Male")),
    
    ethnic = factor(ethnic,
                    levels = 1:4,
                    labels = c("Hispanic","White","Black","Other/Mixed")),
    
    educ  = factor(educ, levels = 1:5),
    
    increl = factor(increl),
    
    sleep_dur = ifelse(sleep_dur >= 90, NA, sleep_dur)
  )


data_mod <- na.omit(data_sub)


str(data_mod)
nrow(data_mod)

```


Categorical covariates were converted to factor variables to ensure correct modeling of non-numeric levels. Abnormal value are recoded as NA.


Step 2. Exploratory Data Analysis (EDA)

```{r}

hist(data_mod$hdl, main="HDL distribution", xlab="HDL")
boxplot(data_mod$hdl, main="Boxplot of HDL")
summary(data_mod$hdl)

par(mfrow=c(2,3))
plot(data_mod$age, data_mod$hdl, main="HDL vs age")
plot(data_mod$bmi, data_mod$hdl, main="HDL vs BMI")
plot(data_mod$rr_sys, data_mod$hdl, main="HDL vs SBP")
plot(data_mod$hg, data_mod$hdl, main="HDL vs hg")
plot(data_mod$cd, data_mod$hdl, main="HDL vs cadmium")

par(mfrow=c(1,3))
boxplot(hdl ~ male, data=data_mod, main="HDL by sex")
boxplot(hdl ~ ethnic, data=data_mod, main="HDL by ethnic group")
boxplot(hdl ~ educ, data=data_mod, main="HDL by education")

cor_matrix <- cor(data_mod[, c("age","bmi","rr_sys","hg","cd")])
cor_matrix

summary(data_mod[, c("age","bmi","rr_sys","hg","cd","sleep_dur")])

```
Exploratory analyses were conducted on the complete-case dataset to evaluate data quality, characterize distributions, and assess whether linear regression was appropriate for modeling HDL.

The distribution of HDL was unimodal with mild right-skewness, and although several higher observations were present, none appeared excessively distant from the main body of the data. Accordingly, no transformation was required for the outcome.

Continuous predictors (age, BMI, systolic blood pressure, hemoglobin, cadmium, and sleep duration) showed plausible ranges and expected population variability. Scatterplots demonstrated roughly linear trends with HDL, most notably a clear negative relationship with BMI. No variable displayed marked curvature or patterns indicating the need for polynomial or spline terms. 

Categorical predictors—including sex, ethnicity, and education—showed visible group differences in HDL, supporting their inclusion in the regression model.

Pairwise correlations among continuous variables were modest (all |r| < 0.4), suggesting no substantial collinearity at this stage. No influential outliers or implausible values remained after data cleaning and recoding.

Overall, the distributional characteristics, linear-appearing relationships, and absence of major anomalies indicated that a standard multiple linear regression model was appropriate for these data.


Step 3. Specification of the Initial Full Model

```{r}

#association evaluation 
model <- lm(hdl ~ ., data = data_mod)  
summary(model)

library(car)
vif(model)

```

The full model explained approximately 25.82% of the variability in HDL levels, 
Among all predictors, sex, BMI, systolic blood pressure, and cigarette consumption remained statistically significant.

For the categorical predictors, Black ethnicity showed a significantly higher HDL level compared with the reference group (Hispanic), whereas the other ethnicity categories did not reach statistical significance.

Education level and income category, included as multi-level factors, did not demonstrate independent associations with HDL.

To evaluate multicollinearity, we examined the Generalized Variance Inflation Factor (GVIF), which accounts for multi-level categorical variables. All GVIF values were well below 2, indicating no evidence of problematic multicollinearity in the model.

Overall, these diagnostic results support the adequacy of the selected covariates.Accordingly, we proceeded with the full set of 13 variables for subsequent model refinement and interpretation.



Step 4. Model Selection - AIC

```{r}

model_full <- lm(hdl ~ age + male + ethnic + educ + increl +
                   bmi + rr_sys + hg + cd +
                   sleep_dur + drnkprd_prv12mo + cigsprd_prv30d + phq2,
                 data = data_mod, na.action = na.omit)

# model_step <- step(model_full, direction = "backward")

out <- capture.output(
  step(model_full, direction = "backward", trace = TRUE))
last_step_index <- tail(grep("^Step:", out), 1)
final_block <- out[last_step_index:length(out)]
cat(final_block, sep = "\n")


```

Using backward stepwise selection based on AIC, the final HDL model retained seven predictors: sex, ethnicity, BMI, systolic blood pressure, blood cadmium, cigarette consumption, and age.

In the final model, age showed only a weak and statistically non-significant association with HDL (p = 0.076). This suggests that the crude relationship between age and HDL was largely explained by other covariates—particularly BMI, blood pressure, smoking, and cadmium exposure—which are themselves correlated with age. After adjusting for these factors, age did not exhibit an independent contribution to HDL variation.

BMI remained the strongest predictor, showing a pronounced negative association with HDL. Sex and ethnicity also had meaningful effects, with males and certain ethnic groups having higher HDL levels compared with their reference categories. Systolic blood pressure and cigarette consumption showed smaller but statistically robust associations: HDL increased modestly with higher systolic blood pressure and decreased with greater cigarette use.

Cadmium, although borderline in statistical significance, improved model fit and was therefore retained. Removing cadmium increased the AIC from –1013.51 to –1012.73, indicating that cadmium contributed incremental explanatory value despite its non-significant p-value.


Step 5. Assessment of Multicollinearity
```{r}

library(car)

model_final <- lm(hdl ~ age + male + ethnic + bmi + rr_sys + cd + cigsprd_prv30d,
                  data = data_mod)

vif(model_final)


```


Step 6. Model Diagnostics and Assumption Checking

The following diagnostics ensure that the linear model is valid, unbiased, and reliable: 

```{r}

par(mfrow = c(3,2))
plot(model_final)

library(lmtest)
dwtest(model_final)

hist(residuals(model_final), 
     breaks = 30,
     main = "Histogram of Residuals",
     xlab = "Residuals")

cook <- cooks.distance(model_final)
plot(cook, pch=20, main="Cook's Distance")
abline(h = 1, col="red")

```

Diagnostic 1: Linearity （Residuals vs Fitted）

Residuals are mostly randomly distributed around zero, with a slight curvature in the smoother line and mild spreading toward the right side. This indicates that the linearity assumption is generally satisfied.
The slight curvature suggests minimal nonlinearity, but it is not severe enough to threaten model validity.


Diagnostic 2: Normality of Residuals （Q-Q Residuals）


Residuals follow the theoretical normal line closely in the center but deviate upward in the right tail due to a few high-valued observations.The central portion suggests approximate normality. Tail deviations indicate the presence of a few outliers, which is common in biomedical data.

Diagnostic 3: Homoscedasticity (Scale–Location Plot)


Most points are evenly scattered, with a slight increase in spread at higher fitted values. Residual variance is mostly constant, though mild heteroscedasticity may exist at the upper range of fitted values. Homoscedasticity is mostly satisfied


Diagnostic 4: Independence of Residuals (Durbin–Watson Test)

The Durbin–Watson statistic indicated no evidence of autocorrelation, which is expected given the cross-sectional structure of the NHANES dataset.The independence assumption is therefore satisfied.


Diagnostic 5: Outliers (Residuals vs Leverage, Cook’s Distance)

Most observations have low leverage. A few points (e.g., ID 3872) show moderately high leverage, but none approach the Cook’s distance threshold.These high-leverage points do not exert disproportionate influence on the regression coefficients. Alos, all Cook’s D values are far below 1. No observation approaches the typical cutoff for concern. The model is stable and not driven by single influential observations.





Step 7. Predictive Performance Evaluation and Interpretation

```{r}

set.seed(123)

n <- nrow(data_mod)
train_index <- sample(1:n, size = 0.8*n)

train <- data_mod[train_index, ]
test  <- data_mod[-train_index, ]

pred_aic <- predict(model_final, newdata = test)
mse_aic_full <- mean((pred_aic - test$hdl)^2)


model_step_train <- lm(
  hdl ~ age + male + ethnic + bmi + rr_sys + cd + cigsprd_prv30d,
  data = train
)


pred_aic <- predict(model_step_train, newdata = test)
mse_aic_train <- mean((pred_aic - test$hdl)^2)

mse_aic_full
mse_aic_train

summary(model_step_train)

```

In our analysis, the mean squared error (MSE) obtained by applying the full-data model to the test set ( 0.1388587) was slightly lower than the MSE obtained using the training-only fitted model (0.1465096).

This small difference is expected: because the full model was trained on the entire dataset—including observations that later appeared in the test set—it produces optimistically biased (overly favorable) predictions.

Importantly, the two MSE values are very close, indicating that the training-only model performs similarly well on unseen data.

This suggests that the final model does not exhibit overfitting and shows good generalizability to new observations.


```{r}
summary(model_final)
```

The final model for predicting HDL levels included the covariates age, sex, ethnicity, BMI, systolic blood pressure (rr_sys), blood cadmium (cd), and recent cigarette consumption (cigsprd_prv30d).

Adjusting for all other predictors:

• Age
Each additional year of age was associated with a 0.002105 mmol/L increase in HDL (β = 0.00217, p = 0.0850 ). This effect was small and not statistically significant, indicating that age has only a modest independent contribution once demographic and lifestyle factors are controlled.

• Sex
Males had 0.155 mmol/L higher HDL compared with females (β = 0.155403, p < 0.001). This was one of the strongest predictors in the model.

• Ethnicity
Relative to the Hispanic reference group:
White individuals: +0.092 mmol/L (p = 0.0813)
Black individuals: +0.243 mmol/L (p < 0.001)
Other/Mixed individuals: +0.136 mmol/L (p = 0.0491)
Ethnicity was therefore an important overall predictor, with the Black group showing a particularly strong positive association with HDL.

• BMI (Body Mass Index)
Each 1-unit increase in BMI was associated with a 0.0248 mmol/L decrease in HDL (β = -0.025, p < 0.000). BMI was the strongest predictor in the model, consistent with well-established evidence linking adiposity with lower HDL.

• Systolic Blood Pressure (rr_sys)
A 1 mmHg increase in systolic blood pressure corresponded to a 0.0026 mmol/L increase in HDL (β = 0.00266, p = 0.0118). Although statistically significant, the effect size was small and likely not clinically meaningful.

• Blood Cadmium (cd)
Each unit increase in blood cadmium was associated with a 0.0033 mmol/L increase in HDL (β = 0.0032, p = 0.0986). The effect was not statistically significant, but AIC model selection retained cadmium because removing it increased AIC, indicating a worse model fit and suggesting cadmium still provides incremental explanatory value.

• Recent Cigarette Consumption (cigsprd_prv30d)
Each additional cigarette smoked in the past 30 days was associated with a 0.0051 mmol/L decrease in HDL (β = -0.005057, p = 0.0108). This represents a meaningful and statistically significant negative effect, consistent with strong epidemiological evidence that smoking lowers HDL.

Overall, BMI, sex, ethnicity, systolic blood pressure, and cigarette consumption contributed to explaining variation in HDL levels, with BMI and sex being the most influential predictors. 

Although age and cadmium were not individually significant, they were retained based on theoretical relevance and AIC model selection.



7.Cancer

a\. Estimate the lifetime prevalence of cancer. Can you also give an interval estimate?

```{r}

library(binom)
library(tidyr)
library(broom)   

cancer <- sum(data$cancer_ever == TRUE, na.rm = TRUE)
tot <- sum(!is.na(data$cancer_ever))


CI_CP <- binom.test(cancer, tot, conf.level = 0.95)
tidy(CI_CP)


```

The estimated lifetime prevalence of cancer in the sample was 8.8%.

Using an exact binomial method, the 95% confidence interval ranged from 8.0% to 9.6%.Using a Wilson score interval, the estimate was virtually identical (95% CI: 8.0% to 9.6%), with a slightly narrower lower bound.

Overall, the lifetime cancer prevalence is approximately 8–10% in this population.



b\. What are the prevalence estimates in those who were exposed to pollutants at work for a longer time period, and in those who weren’t? Is there a significant difference in prevalence between these two subgroups?


```{r}

x1 <- sum(data$cancer_ever[data$hrsworked_prvwk_cat == "< 40"] == TRUE, na.rm = TRUE)
n1 <- sum(!is.na(data$cancer_ever[data$hrsworked_prvwk_cat == "< 40"]))

x2 <- sum(data$cancer_ever[data$hrsworked_prvwk_cat == ">= 40"] == TRUE, na.rm = TRUE)
n2 <- sum(!is.na(data$cancer_ever[data$hrsworked_prvwk_cat == ">= 40"]))

prop_2sample <- prop.test(c(x1, x2), c(n1, n2), conf.level = 0.95, correct = TRUE) 
prop_2sample


```

The lifetime prevalence of cancer was 5.27% among participants working less than 40 hours per week and 4.29% among those working 40 hours or more.
A two-sample test for equality of proportions showed no statistically significant difference between the two groups (χ² = 0.95, p = 0.33).
The 95% confidence interval for the difference in prevalence (–0.96% to 2.93%) includes zero, indicating that the observed difference is compatible with random variation.
Therefore, there is no evidence of a significant association between longer weekly working hours and cancer prevalence in this sample.


c\. Adjust for age in the relationship between lifetime diagnosis of cancer and exposure to pollutants, using the categorized age variable (Note: No information on pollutant exposure was collected from participants aged 80+, so these cannot be included in the analysis). Does the adjustment for age change the picture? Interpret the model coefficients including the intercept.

```{r}

vars_log <- c("cancer_ever", "workpollut", "age_group")
data_mod_log <- data[, vars_1] %>% na.omit()

data_mod_log$workpollut <- factor(data_mod_2$workpollut)
data_mod_log$age_group  <- factor(data_mod_2$age_group)

model_full_log <- glm(
  cancer_ever ~ workpollut + age_group,
  data = data_mod,
  family = binomial
)

summary(model_full_log)

OR <- exp(coef(model_full_log))
CI <- exp(confint(model_full_log))

results <- cbind(OR, CI)
colnames(results) <- c("OR", "2.5%", "97.5%")

round(results, 3)


```

The logistic regression output shows log-odds coefficients, which were converted to odds ratios (OR) for easier interpretation.

Workplace pollutant exposure was not significantly associated with cancer (OR = 1.11, 95% CI 0.87–1.42, p = 0.39).

Age was a strong predictor: compared with the 18–34 age group, participants aged 35–49, 50–64, and 65–79 had 3.35-, 8.91-, and 22-fold higher odds of cancer, respectively (all p < 0.001).



d.Try to find a good model of cancer diagnosis, describe, and interpret it, as you did for HDL.

7 steps:
Step 0. candidate covariates
Step 1. Data Preparation
Step 2. Exploratory Data Analysis (EDA)
Step 3. Specification of the Initial Full Model
Step 4. Model Selection - AIC
Step 5. Assessment of Multicollinearity
Step 6. Model Diagnostics and Assumption Checking
Step 7. Predictive Performance Evaluation and Interpretation



Step 0. candidate covariates
```{r}

library(dplyr)
vars <- c(

  "age",
  "male",
  "ethnic",
  "educ",

  "bmi",
  "cigsprd_prv30d",
  "drnkprd_prv12mo",

  "cd",
  "hg",
  "pb",    

  "srhgnrl",  
  "hivpos"
)


```


Step 1. Data Preparation
```{r}

```


Step 2. Exploratory Data Analysis (EDA)
```{r}

```



Step 3. Specification of the Initial Full Model
```{r}

```




Step 4. Model Selection - AIC
```{r}

```




Step 5. Assessment of Multicollinearity
```{r}

```




Step 6. Model Diagnostics and Assumption Checking
```{r}

```



Step 7. Predictive Performance Evaluation and Interpretation
```{r}

```








