---
title: "Day 3"
output: html_document
---

#Logistic regression

## Assumptions
The relationship between predictors and log-odds is linear. No errors. 
The effect of a continuous variable x is constant for all values of x.
No assumption are made about the distribtuion of covariables, but should not be highly correlated. 
Large sample sizes are required. The more covariables, the larger sample size required. 
Choice of model should always depend on biological or clinical considerations in addition to statistical results. 

## How R does calculation 

Distribution of P (Maximize Likelihood): S-shape curve of x-marker, y-proportion of deaths ranging from 0-1. 
Derivatives should not be used to get Maximize Likelihood, because it might get a regional rather than the global Max. Instead, Iterative process should be used. 

## Assessment of the fitted model

**CI and tests (Wald test, LR test): **LR tests checks if the model increases the likelihood of obtaining the data compared to null model. 
**Goodness of fit of the model:** Chi-square test checks how far the observed p of outcome is away from the expected p of outcome predicted by logistic regression. 
**R square: ** Here, it is referred to as measures of effect size (how good exploratory variables are in predicting the outcome), not goodness of fit. 
**Discrimination**: How well a model distinguishes patients who survive from who die (AUC).Similar to diagnosis. 

**Validation:** How well the results will generalize to an independent data set. External validation is usually not feasible, but internal validation: **bootstrapping** (drawing samples with replacement, model may be developed in bootstrap samples and tested in the original sample), **cross-validation** (training set and testing set, performed over rounds; Brier score is used to estimate the prediction error). 

**Residual Analysis** Deviance Residual determines if individual points are not fitted by the model; Partial residuals

## Variable selection

**Forward selection** Initial model does not include any covariables. It includes covariables that decrease AIC.
**Backward selection** Initial model includes all covariables. It excludes covariables that decrease AIC.
**Stepwise selection** 

## polychotomous scale, continuous variable and interaction
One-degree FP function, Two-degree FP function
Quantitative interaction and Qualitative interaction


# Classification: Random forest and maschine learning
idea of underfitting, overfitting and trade off. 
## Random forest
many decision trees, and the final classification would be the mode of all results from trees. 
other ML methods: naive bayes, decision tree, linear SVM, nearest neighbors



#Exercise 3

## Research Question:
How does the use of hard drug effect asthma among US citizens? 
Confounder: Age, SCS - income, education, work status; smoking and cannabis

```{r}
library(conflicted)
library(tidyverse)
library(magrittr)
library(readr)
library(ggplot2)
library(gtsummary)
library(boot)
library(MASS)
library(broom)

conflicts_prefer(
  dplyr::filter(),
  dplyr::select()
)

```

## Data Exploration
```{r}

# 3.2.1
library(readr)
df <- read_delim("../../data/NHANES1.csv", delim = ";")

# 3.2.2 
ncol(df)
nrow(df)
dim(df)
colnames(df)
head(df)
df %>% map(class)
summary(df)


df %>%
  select(asthma_ever, asthma_now, harddrg_ever) %>%
  tbl_summary()

df %>% 
  ggplot(aes(x = factor(smokstat)))+
  geom_bar() 
  theme_bw()
  
#ggplot(df) +
# geom_bar(aes(x = factor(smokstat))) +
# theme_bw()


ggplot(df) +
  geom_boxplot(aes(x = male, y = age, fill = male)) +
  theme_bw()

df %>% 
  ggplot(aes(x = factor(increl), fill = increl)) +  
  geom_bar() 
  theme_bw() 
  
df %>% 
  ggplot(aes(x = factor(educ), fill = educ)) +  
  geom_bar() 
  theme_bw() 

df %>% 
  ggplot(aes(x = factor(jobstat_lwk), fill = jobstat_lwk ), alpha = 0.2)+ 
  geom_bar() 
  theme_bw() 
  

df %>% 
  ggplot(aes(x = factor(cannab_ever), fill = cannab_ever))+
  geom_bar() 
  theme_bw()

ggplot(df) +
  geom_density(aes(x = bmi, fill = male), alpha = 0.5) +
  theme_bw()

ggplot(df) +
  geom_density(aes(x = weight, fill = male), alpha = 0.5) +
  theme_bw()

ggplot(df) +
  geom_density(aes(x = height, fill = male), alpha = 0.2) +
  theme_bw()


```

Confounder: Age, SCS - income, education, work status; smoking and cannabis
```{r}

# 3.2.3
summary(df) 
df %>%
  is.na() %>%
  colSums() %>%
  sort(decreasing = TRUE)

df %>%
  select(livdis_ever,livdis_now)%>%
  plyr::count()

df %>%
  select(asthma_ever, asthma_now) %>%
  plyr::count()

# Only participants with `livdis_ever == TRUE` have a value for `livdis_now`. For all participants with `livdis_ever == FALSE`, `livdis_now` should be FALSE and not NA. Same for `asthma_now`.


#3.2.5
# my code
df %>% gtsummary::tbl_cross( row = asthma_ever, col = harddrg_ever, missing = "no")

# F.Foster's code : learn how to tbl_cross while rename the levels
df %>%
  select(harddrg_ever,asthma_ever) %>%
  mutate(
    harddrg_ever = factor(harddrg_ever,
        levels = c(TRUE, FALSE),
        labels = c("yes", "no")
        ),
  asthma_ever = factor(asthma_ever,
        levels = c(TRUE, FALSE),
        labels = c("yes","no")
  ) ) %>%
  tbl_cross(
    row = harddrg_ever,
    col = asthma_ever,
    missing = "no"
  )

round((102 / 471) / (459 / 2457), 2) # calculate by hand

install.packages("crosstable")
library(crosstable)

crosstable::crosstable(df,
 cols = harddrg_ever,
 by = asthma_ever,
 showNA = "no",
 percent_pattern = "{n}",
 effect = T
 ) %>%
  crosstable::as_flextable()

```

## Data prparation
```{r}

# 3.3.1 back up
df.old <- df %>%
  set_colnames(paste0(colnames(.), ".old"))

df <- df %>%
  mutate(
    male = factor(male, 
      levels = c (FALSE, TRUE),
      labels  = c("female", "male")
      ), 
    harddrg_ever = factor(harddrg_ever,
      levels = c (FALSE, TRUE),
      labels = c ("never", "ever")
    ),
    asthma_ever = factor(asthma_ever,
      levels = c (FALSE, TRUE),
      labels = c ("never", "ever")
    )
  )

df %>% select(male, harddrg_ever,asthma_ever) %>% map(class)

# compare the old variable with the new variable, check if factor is successful

cbind(df.old, df) %>%
  select(male.old,male) %>%
  plyr:: count()

cbind(df.old, df) %>%
  select(harddrg_ever.old, harddrg_ever) %>%
  plyr::count()

cbind(df.old, df) %>%
  select(asthma_ever.old, asthma_ever) %>%
  plyr::count()

rm(df.old) # remove back up if success



## 3.3.2
# Adjust `asthma_now` based on your result from task 3.2.3. It will be the
# outcome of our regression.

df %>% 
  select(asthma_ever, asthma_now) %>% 
  plyr::count()
# no current asthma should be: 313 + 4244 = 4557
# current asthma should be: 426
# NA should be: 13 + 4 = 17

df.old <- df %>%
  set_colnames(paste0(colnames(.), ".old"))

#  %<>% 原地赋值管道！！！
df %<>% 
  mutate(asthma_now = case_when(
  asthma_now == TRUE ~ 1,
  asthma_now == FALSE ~ 0,
  is.na(asthma_now) & asthma_ever == "never" ~ 0    
))   # here we replace NA using FALSE, because people might only answer asthma_ever but not asthma_now. 

class(df$asthma_now)

cbind(df.old, df) %>%
  select(asthma_now.old, asthma_ever, asthma_now) %>%
  plyr::count()
# check if successfully corrected

df$asthma_now %>%
  plyr::count()


rm(df.old)

## 3.3.3
## Create the PHQ-9 sum score. Also create a standardised version. Check
# afterwards if it worked.

## 原地赋值管道

df %<>%
  rowwise () %>%
  mutate(
    phq_sum = sum(
      phq1,phq2, phq3, phq4, phq5,phq6,phq7,phq8,phq9,
      na.rm = FALSE
    )
  ) %>%
  ungroup() # why?

# check if successful
df %>% 
  select(phq1,phq2, phq3, phq4, phq5,phq6,phq7,phq8,phq9, phq_sum) %>%
  plyr::count() %>%
  view()

df %<>% mutate(
  phq_sum_std = (phq_sum - mean(phq_sum, na.rm = T))/ sd(phq_sum, na.rm = T)
)

df %>%
  select(phq_sum, phq_sum_std) %>%
  plyr :: count() %>%
  head()   # use head () and view () to get an overview

# Check afterwards if it worked. 
mean(df$phq_sum, na.rm = TRUE) # 3.155804
sd(df$phq_sum, na.rm = TRUE)  # 4.419932



```

## Modeling

```{r}
# 3.4.1
# Fit a logistic regression for outcome `asthma_now`. As your initial model,
# fit the intercept-only model. How many individuals were excluded from the
# analysis? Calculate the proportion of participants with `asthma_now` from the
# estimated parameter.

m0 <- glm(asthma_now ~ 1, data = data, family = binomial(link = "logit"))
nrow(df) - nobs(m0) 
summary(m0)  

boot::inv.logit(m0$coefficients[1]) 
inv.logit(coef(m0))
inv.logit(confint(m0))



# 3.4.2
m1 <- glm(asthma_now ~ harddrg_ever, data = df, family = binomial(link = "logit"))
nobs(m1) # 3481

m1 %>% gtsummary::tbl_regression(exponentiate = T)
# tbl_regression(m1, exponentiate = T)


df %<>%
  mutate(smoke = case_when(
    smokstat == 1 ~ "non-smoker",
    smokstat == 2 ~ "smoker",
    smokstat == 3 ~ "smoker"
  ),
  smoke = as.factor(smoke)
  )

# check if successur

class(df$smoke)

df %>%
  select(smokstat,smoke) %>%
  plyr::count()


m2 <- glm(asthma_now ~ harddrg_ever + male + age + smoke, data = df, family = binomial(link = "logit"))

gtsummary::tbl_regression(m2,exponentiate = T)
nobs(m2)
tbl_regression(m2, exponentiate = T)

#3.5.1
AIC(m0,m1,m2)

```


```{r}

## 3.5.2
# Split the potential confounders, for which you were not certain that they are
# potential confounders, into variable sets. Continue to modify your model by
# including these variable sets as predictors. Follow the workflow: fit,
# evaluate, compare, modify.


m3 <- glm(asthma_now ~ harddrg_ever + male + age + smoke + phq_sum_std,
  data = df, family = binomial(link = "logit")
)
summary(m3)
nobs(m3)
tbl_regression(m3, exponentiate = T)

AIC(m2, m3)


## 3.5.3
# Compare point and interval estimate of `harddrg_ever` in a plot using
# `geom_pointrange()`.

est_m1 <- m1 %>%
  tidy(conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(
    type = "Crude"
  ) %>%
  filter(term == "harddrg_everever")

est_m2 <- m2 %>%
  tidy(conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(
    type = "Adjusted (part)"
  ) %>%
  filter(term == "harddrg_everever")

est_m3 <- m3 %>%
  tidy(conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(
    type = "Adjusted (full)"
  ) %>%
  filter(term == "harddrg_everever")

rbind(est_m1, est_m2, est_m3) %>%
  mutate(type = as.factor(type)) %>%
  ggplot() +
  geom_pointrange(aes(
    x = estimate,
    xmin = conf.low,
    xmax = conf.high,
    y = type
  )) +
  geom_vline(aes(xintercept = 1), linetype = "dashed", color = "grey") +
  theme_bw() +
  theme_sub_axis_y(title = element_blank()) +
  labs(x = "OR")

# 3.5.4: models are not all fitted to the same number of observations


```




















