---
title: "Day4"
output: html_document
---


**Research Questions:**

Is age related with having a history of cancer among American citizens?
Is smoking related with having a history of cancer among American citizens?

```{r}
# 4.1

library(conflicted)
library(tidyverse)
library(magrittr)
library(readr)
library(ggplot2)
library(gtsummary)
library(boot)
library(MASS)
library(broom)

conflicts_prefer(
  dplyr::filter(),
  dplyr::select()
)

```

# Logistic Regression

```{r}

library(readr)
# df <- read_delim("../../data/NHANES1.csv", delim = ";") # had problem, so switch to read_csv2

df <- read_csv2("../../data/NHANES1.csv")

```


## Data Exploration
```{r}
df %>%
  select(age, smokstat, cancer_ever) %>%
  tbl_summary()

df %>% 
  ggplot(aes(x = factor(smokstat), y = age))+
  geom_boxplot() 
  theme_bw()
  
df %>% ggplot() +
  geom_boxplot(aes(x = factor(cancer_ever), y = age, fill = cancer_ever)) +
  theme_bw()

# method 1
df %>% gtsummary::tbl_cross(row = cancer_ever, col = smokstat, missing = "no")


# method 2
library(crosstable)

crosstable::crosstable(df,
 cols = smokstat,
 by = cancer_ever,
 showNA = "no",
 percent_pattern = "{n}",
 effect = T
 ) %>%
  crosstable::as_flextable()

# here we have OR < 1, could be reverse causation or survival bias.

```

## Data Exploration
```{r}

#4.3.1
#Dichotomise variable `smokstat` with categories „non-smoker“ and „smoker“.Transform it into a factor. Check afterwards if it worked.


df %>% 
  select(cigsprd_prv30d, smokstat) %>% 
  plyr::count()

# 1 = no smoker and 2 = occational smokers, 3 = smokers
df <- df %>%
  mutate(smoke_bin2 = case_when(
    smokstat == 1 ~ "non-smoker",
    smokstat == 2 ~ "smoker",
    smokstat == 3 ~ "smoker",
    !is.na(cigsprd_prv30d) & is.na(smokstat) ~ "smoker" 
    # different from the solution, I added this line
  ),
  smoke_bin2  = as.factor(smoke_bin2)
  )

class(df$smoke_bin2) # but i feel this is not always the right way to check if it worked. 


df %>% 
  select(cigsprd_prv30d, smokstat,smoke_bin2) %>% 
  plyr::count()


# 4.3.2
# Dichotomise variable `age` with categories „<60 years“ and „≥60 years“. Transform it into a factor. Check afterwards if it worked.

df <- df %>%
  mutate(age_bin2 = case_when(
    age < 60 ~ "<60",
    age >= 60 ~ ">=60"
  ),
  age_bin2 = factor(age_bin2,
                    levels = c ("<60",">=60" ) 
  )) 
# pay attention to the order of levels, otherwise different baseline group
# levels are different from labels; once add labels, the value will be labels
df %>%
  select(age, age_bin2) %>%
  plyr::count()

## 4.3.3
# From the dichotomised versions of `smokstat` and `age`, create a variable, that encodes their four different combinations, i.e., smoking+young, smoking+old, etc. Turn it into a factor. Check afterwards if it worked.

df <- df %>%
  mutate(smokstat_age = case_when(
    smoke_bin2 == "non-smoker"& age_bin2 == "<60" ~ "young non-smoker",
    smoke_bin2 == "non-smoker"& age_bin2 == ">=60" ~ "old non-smoker",
    smoke_bin2 == "smoker" & age_bin2 == "<60" ~ "young smoker",
    smoke_bin2 == "smoker"& age_bin2 == ">=60" ~ "old smoker"
  ),
  smokstat_age = factor(smokstat_age,
        levels = c(
          "young non-smoker","old non-smoker", "young smoker","old smoker"))
  )

# level and label

df %>% select(smoke_bin2, age_bin2,smokstat_age) %>%
  plyr::count()


```

## Modeling - Initiation
```{r}

## 4.4.1
# Fit and evaluate a logistic regression model for outcome `cancer_ever` and the dichotomised versions of `smokstat` and `age` as predictors. Interpret the model coefficients and write down your answer.

m1 <- glm(cancer_ever ~ smoke_bin2 + age_bin2, 
          data = df, family = binomial(link="logit"))

summary(m1)
m1 %>% gtsummary::tbl_regression(exponentiate = T)
# in the output the outcome is separated, so use predict() to display all OR

m1$xlevels$age_bin2  # check if the value of this variable in model
predict(m1,
  newdata = data.frame(
    smoke_bin2 = c("non-smoker", "non-smoker", "smoker", "smoker"),
    age_bin2 = c("<60", ">=60", "<60", ">=60")
  ),
  type = "response" # output is p - probability
)

# Interpretation:
# Individuals who are 60 years or older have a 4.6 times higher odds of ever having cancer compared to individuals below 60 years of age. This translates into a 22.9% vs. 6.1% probability in non-smokers and a 14.1% vs. 3.4% probability in smokers. 

```

## Modeling 
```{r}

m2 <- glm(cancer_ever ~ smoke_bin2* age_bin2, data = df, family = binomial(link="logit"))

# m2 <- glm(cancer_ever ~ smoke_bin2 + age_bin2 + smoke_bin2* age_bin2, data = df, family = binomial(link="logit"))，does not influence the outcome but not recommended

summary(m2)
nobs(m2)
m2 %>% gtsummary::tbl_regression(exponentiate = T)

newdata_m2 <-data.frame(
    smoke_bin2 = c("non-smoker", "non-smoker", "smoker", "smoker"),
    age_bin2 = c("<60", ">=60", "<60", ">=60"))

newdata_m2$logodds <-predict(m2,newdata = newdata_m2,type = "link") # create logodds
newdata_m2$odds <- exp(predict(m2,newdata = newdata_m2,type = "link"))
newdata_m2$prob <-predict(m2,newdata = newdata_m2,type = "response")

print(newdata_m2)

# calculate OR

or_smoke_in_young <- round(newdata_m2$odds[newdata_m2$smoke_bin2 == "smoker" & newdata_m2$age_bin2 == "<60"]/newdata_m2$odds[newdata_m2$smoke_bin2 == "non-smoker" & newdata_m2$age_bin2 == "<60"],2)


or_smoke_in_old <- round(newdata_m2$odds[newdata_m2$smoke_bin2 == "smoker" & newdata_m2$age_bin2 == ">=60"]/newdata_m2$odds[newdata_m2$smoke_bin2 == "non-smoker" & newdata_m2$age_bin2 == ">=60"],2)

or_age_in_smoker <- round(newdata_m2$odds[newdata_m2$smoke_bin2 == "smoker" & newdata_m2$age_bin2 == "<60"]/newdata_m2$odds[newdata_m2$smoke_bin2 == "smoker" & newdata_m2$age_bin2 == ">=60"],2)

or_age_in_nonsmoker <- round(newdata_m2$odds[newdata_m2$smoke_bin2 == "non-smoker" & newdata_m2$age_bin2 == "<60"]/newdata_m2$odds[newdata_m2$smoke_bin2 == "non-smoker" & newdata_m2$age_bin2 == ">=60"],2)

cbind(or_smoke_in_young, or_smoke_in_old, or_age_in_nonsmoker,or_age_in_smoker) 

```

## Modeling-Evaluation
```{r}
## 4.4.3
# Look for a R package which provides the option to plot the coeffecients of multiple regression models side by side. Apply it to create a basic comparison of the previous two models.

#install.packages("jtools")
#install.packages("sjPlot")
install.packages("visreg")


jtools::plot_summs(m1,m2)
sjPlot::plot_models(m1,m2)


library("visreg")
visreg(m1, scale = "response")
# m1 without interaction, one image for each variable

visreg(m2, "smoke_bin2", by = "age_bin2")
# output is log-odds


```

```{r}

## 4.5.1
# Modify your model and only use the combined variable of the dichotomised versions of `smokstat` and `age`, i.e., with categories smoking+young, smoking+old, etc., as predictor. Fit and evaluate the model. Compare it with
# the interaction model.

m3 <- glm(cancer_ever ~ smokstat_age, data = df, family = binomial(link = "logit"))

m3 %>% gtsummary::tbl_regression(exponentiate = T)



newdata_m3 <- data.frame(smokstat_age = c("young non-smoker", "old non-smoker","young smoker", "old smoker"))

newdata_m3$logodds <- round(predict(m3, newdata = newdata_m3, type = "link"),2)
newdata_m3$odds <- round(exp(predict(m3, newdata = newdata_m3, type = "link")),2)
newdata_m3$prob <- round(predict(m3, newdata = newdata_m3, type = "response"),2)

print(newdata_m3)
print(newdata_m2) # identical results


```


```{r}

m4 <- glm(cancer_ever ~ smoke_bin2 * age, data = df, family = binomial(link = "logit"))
gtsummary::tbl_regression(m4, exponentiate = TRUE)


newdata_m4 <- data.frame(
  age = c(70,25),
  smoke_bin2 = c ("non-smoker","smoker")
)

newdata_m4$odds <- exp(predict (m4, newdata = newdata_m4, type = "link" ))
newdata_m4$prob <- predict(m4, newdata = newdata_m4, type = "response" )

print(newdata_m4)

# the use of predict (), exp(), data.frame()


```

# Generalized Additive Models
Research Question: What is the relationship between weight and HDL?


## Data Exploration
```{r}

# 4.8
# Explore the distributions of HDL and weight. Also plot their relationship, including “lm” and “loess” smooths. Use `geom_smooth()` with the `method` argument.


df %>% 
  select(hdl, weight) %>%
  summary()

df %>% 
  ggplot()+
  geom_density(aes(x = hdl), color = "blue") +
  geom_density(aes(x = log(hdl)), color = "red") +
  theme_bw()
  
# combine two images
df %>% 
  ggplot(aes(x = weight, y = hdl))+
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, color = "red") +
  geom_smooth(method = loess, color = "blue") 
  theme_bw()

```

## Data Preparation

```{r}
# 4.9
# Create the logarithm of HDL as new variable. Check afterwards if it worked.


df <- df %>% mutate(hdl_log = log(hdl))

df %>% 
  select(hdl_log, hdl) %>% 
  plyr::count() %>% 
  head()
# continuous variable

```

## Modeling

```{r}
## 4.10.1
# initial model is linear and uses the non-logarithmic version of HDL. Fit and evaluate.

lm_mod <- lm(hdl ~ weight, data = df)

summary(lm_mod)                                       #output 1
tidy(lm_mod, conf.int = TRUE)                         #output 2
gtsummary::tbl_regression(lm_mod)                     #output 3
gtsummary::tbl_regression(lm_mod,exponentiate = TRUE) #output 4


visreg(lm_mod, gg = T, line = list(col="green")) +
  stat_smooth(data = df, aes(x = weight, y = hdl_log), method = "loess", col = "red")
  theme_classic()
  
  
plot(lm_mod)  # assumption check

lm_mod %>% 
  simulate(nsim = 100) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  ) %>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") +
  geom_density(
    data = df,
    mapping = aes(x = hdl),
    linewidth = 1
  ) +
  theme_bw()

# the distribution of outcome is skewed, consider log transformation
ggplot() + geom_density(data = df,mapping = aes(x = hdl))

```



```{r}
## 4.10.2
# Modify your model to include log(HDL) as outcome. Fit, evaluate, and compare.

lm_mod_log <- lm(hdl_log ~ weight, data = df)

visreg(lm_mod_log, gg = T, line = list(col="red")) +
  stat_smooth(data = df, aes(x = weight, y = hdl_log), method = "loess", col = "green")
  theme_classic()
  
# if method = "lm", the line will be exactly line = list(col = "red") 
  
```


```{r}


## 4.10.3
# Modify your model to include a non-linear relationship with weight. Fit, evaluate, and compare the model. Look at the summary. How wiggly do you expect the smooth to be? Also plot the non-linear relationship (e.g., with package visreg).

# install.packages("mgcv")
library(mgcv)

gam_mod_log <- gam(hdl_log ~ s(weight), data = df, family = gaussian())
summary(gam_mod_log)
# edf: 3.7 => relatively wiggly, but not extremely wiggly

visreg(gam_mod_log, gg = T, line = list(col="red")) +
  stat_smooth(data = df, aes(x = weight, y = hdl_log), method = "loess", col = "green")
  theme_classic()

gam_mod_log %>% 
  simulate(nsim = 100) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  ) %>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") +
  geom_density(
    data = df,
    mapping = aes(x = hdl_log),
    linewidth = 1
  ) +
  theme_bw()


AIC(lm_mod,lm_mod_log,gam_mod_log)

```


```{r}
# 4.10.4
# Extract smoothing parameter and number of basis functions from your model.Try out different values for `sp` and `k` and see how the smooth changes.

gam_mod_log$sp # smoothing parameter
gam_mod_log$smooth[[1]]$bs.dim  # this is knobs , the k

gam.check(gam_mod_log)
# check if we need to adjust parameter after gam
# The basis dimension check did not indicate undersmoothing (k-index = 1, p = 0.46), and the estimated edf (3.74) is well below k′ (9), suggesting that the chosen k is adequate.
```

```{r}


gam_mod_log2a <- gam(hdl_log ~ s(weight, k= 3), data = df, family = gaussian())

gam_mod_log2b <- gam(hdl_log ~ s(weight, k= 30), data = df, family = gaussian())

gam_mod_log2c <- gam(hdl_log ~ s(weight), sp=0.01, data = df, family = gaussian())

gam_mod_log2d <- gam(hdl_log ~ s(weight), data = df, sp = 100)

visreg(gam_mod_log2a, gg = T) + theme_bw()

visreg(gam_mod_log2b, gg = T) + theme_bw()
# With high k, sp is larger to smooth it back to the optimal shape

visreg(gam_mod_log2c, gg = T) + theme_bw()
# With low sp, edf are higher ==> counter-acted by increased wiggliness

visreg(gam_mod_log2d, gg = T) + theme_bw()
# With high sp, the smooth is forced towards a straight line

```


```{r}

# 4.10.5:  Fit a GAM for outcome `cancer_ever` with predictors smoking and age. Then modify your model to include different smooths of age for smokers and non-smokers. Plot your results.

gam3 <- gam(cancer_ever ~ smoke_bin2 + s(age), data = df, family = binomial(link = "logit"))

# family = Gaussian when the y is continuous, family = binomial(link = "logit") when y is binary; when family = binomial, glm = 直线版 logistic；gam = 曲线版 logistic

visreg(gam_mod3, gg = T)

# gam4 <- gam(cancer_ever ~ s(age, by = smoke_bin2), data = df, family = binomial(link = "logit"))

gam4 <- gam(cancer_ever ~ smoke_bin2 + s(age, by = smoke_bin2), data = df, family = binomial())

visreg(gam4, xvar = "age", by = "smoke_bin2", overlay = T)




```
“In the plot, the y-axis shows the smooth effect of age on the log-odds of ever having cancer (f(age)), with separate smooth curves for smokers (blue) and non-smokers (red); the shaded bands represent uncertainty (≈95% CI), and a separation or crossing of the curves indicates effect modification.

The association between age and cancer prevalence differs by smoking status, indicating effect modification. In smokers the age effect appears weaker at older ages, whereas in non-smokers the predicted risk continues to increase, with the curves crossing around age ~60.