---
title: "Day 2"
output: html_document
---

**statistic model:**
* Describe data-generating process, makes assumption about the type of variable and the shape of distribution (e.g., normally distributed)

* Assumption about the distribution of the outcome: 
** Linear regression - normal distributed outcome; if not, the fit is not good. Example: blood pressure, cholesterol, Iron in blood, weight

* simple check :
```{r}
model <- lm(cty ~ 0 + cyl, data = mpg) 

nobs(model) # to check the number of observations
coef(model)
confint(model)
fitted(model) # get fitted value 

resid(model) # get residuals 
rstandard(model)

# broom and tidy
install.packages("broom")
tidy(model, conf.int = TRUE) # reate a tidy output
model %>% tbl_regression() %>% add_n()


# predict values for new observations predict (),  
# use visreg to visualize the parameter estimates 
newdata <- data.frame(cyl = c(100,200)) 
predict (model, newdata = newdata, se.fit = TRUE, interval = "confidence", level = 0.95)

```

* Diagnostics of model
** 1: errors are distributed normally
** 2: error variance is constant
** 3: error for one observation is not influenced by the value of any other error
** 4: individual outlines might strongly influence the parameter estimates
If assumptions are violated? e.g., can increase the complexity

```{r}

plot(model)

# Q-Q plots: Check assumption 1. We compare the residuals to standard normal-Normal (0,1), we want dots on the line. 
# Residuals vs Fitted: Check assumption 2. We want the red line to be horizontal. 
# Scale-Location: We want the red line to be horizontal. 
# Residuals vs leverage : check assumption 4. residuals = outcome outlines. leverage = predictor outlines; cook's distance (the curves). We want the red line to be horizontal. 


```

*model misspefification
If we use a model to simulate data, the model should look like the observed data. 
All the plots mentioned above, look at model misspecification from different angels. 

```{r}
# 1 do the simulation first
simulate(model, nsim = 5, seed = 2175)
# nsim = 5, simulate 5 times
# seed = by setting the same starting value, they are integers, do no mean anything. 

# 2 plot the simulation
# 3 compare with the real data. Ideally, the real data is within the simulation; else, the model doesn't fit very well. By including more predictors, the fit will be improved

```

*compare fit
```{r}

model0 <- lm(cty ~ 1, data = mpg) 
extractAIC(model0)
extractAIC(model)
AIC(model0, model) # AIC should be used to compare models only when the models use the same dataset.
anova(model0, model) # Used for nested models, model fit is tested against each other

# ? if the missingness is different in the two model. Solution? 
# some people speak against stepwise AIC, should be based on domain knowledge. 


```

**LIVE CODING** 

0 Package preparation
```{r}
?MASS::hills # loading packages

install.packages("visreg")
install.packages("gglm")

library(conflicted)
library(tidyverse) # we can check core tidyverse packages by ?
library(MASS)
library(magrittr)
library(gtsummary)
library(broom)
library(gglm)
library(visreg) # to visualise regression model

conflicts_prefer(dplyr::filter(), dplyr::select())
conflict_scout()

```

1 Research question: 
How distance (longer or shorter) affect time of race?

2 Data Exploration
```{r}

dim(hills) # dimensions
head(hills) # a brief summary

hills %>% map(class)
hills %>% lapply(class)
hills %>% 
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    missing = "always"
  )

hills %>% 
  ggplot(aes(x = time)) +
  geom_density() +
  xlim(0, 250)  # from 0 to 250, the x aex
  theme_bw()

hills %>% 
  ggplot(aes(x = dist)) +
  geom_density() +
  xlim(0, 30) 
  theme_bw()
  
hills %>% 
  ggplot(aes(x = climb)) +
  geom_density() +
  theme_bw()


hills %>% 
  ggplot(aes(x = dist, y = time)) +
  geom_point() +
  theme_bw()

hills %>% 
  ggplot(aes(x = climb, y = time)) +
  geom_point() +
  theme_bw()


```

3 Data Preparation: creat variables, and check if it is sucessfully created
```{r}

one_mile_equals_km <- 1.60934
one_foot_equals_m <- 0.30480

data <- hills %>%
  mutate(
    dist_km = dist* one_mile_equals_km,
    climb_m = climb *one_foot_equals_m,
    climb_m_bin = case_when(
      climb_m < 500 ~ "flat",
      climb_m >= 500 ~ "steep"
    ),
    climb_m_bin = as.factor(climb_m_bin)
  ) %>%
  rownames_to_column("nameeee") # assign a name to the colume

rm(one_mile_equals_km,one_foot_equals_m) # remove, and make the environment clean

data %>%
  select(dist, dist_km) %>%
  plyr::count() %>%
  view()

data %>%
  select(climb, climb_m) %>%
  plyr::count() %>%
  view()

summary(data$climb_m_bin)
class(data$climb_m_bin)

data %>%
  select(climb_m, climb_m_bin) %>%
  plyr::count() %>%
  view()

data %>%
  ggplot(aes(x = dist_km)) +
  geom_boxplot() +
  xlim(0, 50) 
  theme_bw()

```

4 modeling
```{r}

# fit
m0 <- lm(time ~ 1, data = data)
nobs(m0)
summary(m0) # intercept here is the mean of time

#evaluate
tidy(m0,conf.int = TRUE)
coef(m0)
confint(m0)

fitted(m0) # here all values are the same, because that's mean
resid(m0)
data$time - fitted(m0)
data$time - resid(m0) # observed value - residule = our intercept, which is mean here. 
plot(m0) # obviously, it is a bad model.


# + dist_km
m1 <- lm(time ~ dist_km, data = data)
nobs(m1)
summary(m1)

m1 %>% tbl_regression(intercept = TRUE)


m1 %>% visreg("dist_km")

simulate(m1, nsim = 100) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  )%>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") +
  geom_density(data = data, mapping = aes(x = time), linewidth = 1) +
  theme_bw()

# + dist_km and climb_m

m2 <- lm(time ~ dist_km + climb_m, data = data)
nobs(m2)
summary(m2)

m2 %>% tbl_regression(intercept = TRUE)


simulate(m2, nsim = 100) %>%   # better fit compared to m1 and m0
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  )%>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") +
  geom_density(data = data, mapping = aes(x = time), linewidth = 1) +
  theme_bw()

AIC(m0,m1,m2) # m2 is the best
anova(m0,m1,m2) # same result

# + dist_km * climb_m 

m3_0 <- lm(time ~ dist_km * climb_m_bin , data = data) 
# the output reflects the coefficents of all terms
# m3 <- lm(time ~ dist_km + climb_m + dist_km * climb_m_bin , data = data)

m3_0 %>%
  tidy(conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high)

intercept <- 17.030822
b_dist <- 1.875182
b_climb <- -13.919384
b_dist_climb <- 3.100137

intercept + b_dist *5 + b_climb * 500 + b_dist_climb * 5*500 
intercept + 5*(b_dist + b_dist_climb * 500) + b_climb * 500
intercept + 500* (b_climb + b_dist_climb * 5) + b_dist * 5
# this illustrates how to understand and interprate our interaction term 


```

======== Exercise ==========


```{r}

m3 <- lm(time ~ dist_km * climb_m, data = data)
summary(m3)


intercept_1 <- 9.3954374
b_dist_km <- 1.875182
b_climb_m <- -0.0031857
b_dist_climb_1 <- 0.0020041

intercept_1 + b_dist_km *5 + b_climb_m  * 500 + b_dist_climb_1 * 5*100 
intercept_1 + b_dist_km *5 + b_climb_m  * 1000 + b_dist_climb_1 * 5*1000
intercept_1 + b_dist_km *10 + b_climb_m  * 500 + b_dist_climb_1 * 10*100
intercept_1 + b_dist_km *10 + b_climb_m  * 1000 + b_dist_climb_1 * 10*1000 


```

1 Research Question: 
Does the relationship between heart weights and body weight in adult cats differ between sex?

2 Data Exploration 
```{r}

?MASS::cats

head(cats)

dim(cats)  # check dimension
cats %>% map(class)  # check class

cats %>%    # check missing value and summary
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    missing = "always"
  ). 


# geom_bar(), geom_density(), geom_histogram()
# Sex - factor, Bwt - numeric, Hwt - numeric

cats %>% 
  ggplot(aes(x = Sex)) + 
  geom_bar() +
  theme_bw() 

cats %>% 
  ggplot(aes(x = Bwt)) + 
  geom_density() +
  xlim(0, 5) +
  theme_bw()

cats %>% 
  ggplot(aes(x = Hwt)) + 
  geom_density() +
  xlim(0, 25) +
  theme_bw()

# when the parameter is a variable, put it in the aes(), if not, outside aes (); 
cats %>% 
  ggplot(aes(x = Bwt, y = Hwt, shape = Sex)) + 
  geom_point(color = "blue") +    # for color or shape that is not a variable, try put in geom_point;
  theme_bw()


cats %>% 
  ggplot(aes(x = Bwt, y = Hwt, shape = Sex,color = Sex)) + 
  geom_point() +    # for color or shape that is not a variable, try put in geom_point;
  theme_bw()
```


3 Data Preparation
```{r}

exe_data <- cats %>%
  mutate(
    female = case_when(
      Sex == "F" ~ 1,
      Sex == "M" ~ 0
    ),
    Bwt_sd = (Bwt - mean (Bwt)) / sd (Bwt),   # learn the implanted function mean() and sd ()
    Hwt_sd = (Hwt - mean (Hwt)) / sd (Hwt)
  )


```

4 Modeling
```{r}
# modeling 1
exe_model0 <- lm(Hwt ~ 1, data = exe_data) 
nobs(exe_model0)


exe_model1 <- lm(Hwt ~ Bwt, data = exe_data) 

tidy(exe_model1,conf.int = TRUE)
# exe_model1 %>% tidy(conf.int = TRUE)
exe_model1 %>% tbl_regression()

summary(exe_model1)
coef(exe_model1)
confint(exe_model1)

exe_data %>% 
  ggplot(aes(x = Bwt, y = Hwt, shape = Sex,color = Sex)) + 
  geom_point() +    # for color or shape that is not a variable, try put in geom_point;
  theme_bw() +
  #geom_abline(intercept = -0.3567 , slope = 4.0341)   
  geom_abline(intercept = exe_model1$coefficients[1], slope = exe_model1$coefficients[2])


```

```{r}

# modeling 2

fitted_Hwt <- predict (exe_model1, newdata = exe_data, se.fit = TRUE, interval = "confidence", level = 0.95)

fitted_Hwt$fit. # in this part, the first colume is what we want
exe_data <- exe_data  %>% mutate(
  fit_Hwt = fitted_Hwt$fit[,1]  # first colume 
)  # fitted(exe_model1) is an alternative


newdata <- data.frame(Bwt = c(5,6,7)) 
predict (exe_model1, newdata = newdata, se.fit = TRUE, interval = "confidence", level = 0.95)


df_fitted_resid <- data_frame(var1 = fitted(exe_model1),var2= resid(exe_model1))
df_fitted_resid

plot(exe_model1)

exe_model2 <- lm(Hwt ~ Bwt + Sex, data = exe_data) 
tidy(exe_model2,conf.int = TRUE)
AIC(exe_model1,exe_model2)
anova(exe_model1,exe_model2)

exe_model3 <- lm(Hwt ~ Bwt * Sex, data = exe_data) 
tidy(exe_model3,conf.int = TRUE)

AIC(exe_model1,exe_model2,exe_model3)
anova(exe_model3,exe_model1,exe_model2) # same conclusion drawn
anova(exe_model1,exe_model2,exe_model3) 


```

```{r}
plot(exe_model1)
plot(exe_model2)
plot(exe_model3)
```






