---
title: "Day 7 Survival Analysis and Cox regression"
output: html_document
---

*The standard error of KM is higher on the left-hand side of the curve because of smaller sample size. 

*Survival distribution of different groups can be compared with the log-rank test. 

*We do not only look at the risk of death at one specific time point (e.g., 1-year survival), but at all time points. Because, otherwise we will lose all the information. 

* Survival function is 1-function of death (failure function). 

* We look at medien survival but not mean survival, because mean is subjective to outliers (even if there is no censoring)

* Outcome of interest = Time-to-event (measured in years, months, days, hours, etc.)

* Time-to-event is numeric, positive, random variable. Normal distribution is less used because time-to-event is skewed. 

*Right censoring, left censoring, interval censoring. Left cencoring may occur when we do not diagnose the cancer when it is there.

* Summarising KM: 1-year survival, 5-year survival, median survival, Convidence Interval, of S(t) and medien survival 

* The log-rank test is based on chi-square test: add together over all event times.


```{r}
library(survival)
library(survminer)
library(ggplot2)
library(MASS)
library(gtsummary)
library(dplyr)
library(conflicted)
library(readr)

conflicts_prefer(
  dplyr::select(),
  dplyr::filter()
)
conflict_scout() # check the conflicts

install.packages("autoReg")


d <- autoReg::anderson
d %>% tbl_summary()
class(d$time)
view(d)
# no missing value, no censoring, only alive and dead at the end of observation.

d%<>% mutate(time = death - diag)
d$time <- as.numeric(d$time)


# overall survival 
time_to_event <- Surv(time = d$time, event = d$status) # a few censored with +
km <- survfit(time_to_event ~ 1, data = d)
summary(km)
print(km)

km_plot <- ggsurvplot(km,
                      data = d,
                      linewidth = 0.5,
                      surv.median.line = "hv",
                      risk.table = TRUE,
                      risk.table.height = 0.3,
                      pval = TRUE,
                      conf.int = TRUE,
                      palette = c("#E69F00"),
                      xlab = "Time(days)",
                      ylab = "Survival probability",
                      title = "KM Curve",
                      ggtheme = theme_minimal())
km_plot


# survival by sex
km_by_sex <- survfit(time_to_event ~ sex, data = d)
summary(km_by_sex)

survdiff(Surv(time, status) ~ sex, data = d)

km_plot_by_sex <- ggsurvplot(km_by_sex,
           data = d,
           risk.table = TRUE, 
           risk.table.height = 0.3,
           pval = TRUE, 
           conf.int = TRUE,        
           legend.labs = c("Female", "Male"),
           legend.title = "Sex",
           palette = c("#E69F00","#56B4E9"), 
           surv_median.line = "hv" , 
           xlab = "Time (days)",
           ylab = "Survival probability",
           title = "Kaplan-Meier Survival Curve by Sex",
           ggtheme = theme_minimal()) 

km_plot_by_sex

coxph(Surv(time, status) ~ sex, data=d) # HR = exp (coef) = 0.7333, the male has 0.733 decreased risk of dying. 


#7.1.5

km <- survfit(time_to_event ~ 1, data = d)
# km$n.cercor is under km

d1 <- data.frame(event_time = km$time,
                 n_risk = km$n.risk, # people at risk
                 n_death = km$n.event, # number of death
                 n_remission = km$n.censor, # number of remission
                 cum_survival = km$surv
                 )
d1 %<>% mutate(prob_death_noncum = n_death/n_risk,
              prob_surv_point = 1-n_death/n_risk)

d1

cumprod(d1$prob_surv_point) # verified.


```

# Data nepalDHS
```{r}

d2 <- read_delim("../../data/nepalDHS.csv", delim = ",") 
d2 %>% tbl_summary()

# trap! recode the variable Alive
d2 %<>% mutate(Status_Alive = case_when(Alive == 1 ~ 0,
                                       Alive == 0 ~ 1))

time_to_event_2 <- Surv(time = d2$time, event = d2$Status_Alive)

km_nepal <- survfit(time_to_event_2 ~ 1, data = d2 )

d2 %>% ggplot() + geom_histogram(aes(x=time))


km_nepal_plot <- ggsurvplot(km_nepal,
                      data = d2,
                      linewidth = 0.5,
                      surv.median.line = "hv",
                      risk.table = TRUE,
                      risk.table.height = 0.3,
                      pval = TRUE,
                      conf.int = TRUE,
                      palette = c("#E69"),
                      xlab = "Time(days)",
                      ylab = "Survival probability",
                      title = "KM Curve",
                      ggtheme = theme_minimal())
km_nepal_plot


# break by Urban
km_nepal_Urban <- survfit(time_to_event_2 ~ Urban, data = d2 )
summary(km_nepal_Urban)

km_nepal_plot_Urban <- ggsurvplot(km_nepal_Urban,
           data = d2,
           risk.table = TRUE, 
           risk.table.height = 0.4,
           pval = TRUE, 
           conf.int = TRUE,        
           legend.labs = c("Female", "Male"),
           legend.title = "Sex",
           palette = c("#E69F00","#56B4E9"), 
           surv_median.line = "hv" , 
           xlab = "Time (days)",
           ylab = "Survival probability",
           title = "Kaplan-Meier Survival Curve by Sex",
           ggtheme = theme_minimal()) 

km_nepal_plot_Urban



```

```{r}
# 7.2.4: For the survival model with `FirstInfantDeath`

# survival rates at 6 months after birth,
# survival rates at 1 year after birth for each group.

km_nepal_FirstInfantDeath <- survfit(time_to_event_2 ~ FirstInfantDeath, data = d2 )

summary(km_nepal_FirstInfantDeath)
# extract the median times


km_nepal_FirstInfantDeath

tbl_merge(list(
  
))

```

