---
title: "Day 3"
output: html_document
---

#Logistic regression
## Assumptions
no assumption are made about the distribtuion of covariables, but should not be highly correlated. 
Large sample sizes are required. The more covariables, the larger sampl size required. 
Choice of model should always depend on biological or clinical considerations in addition to statistical results. 

## How R does calculation 

Distribution of P (Maximize Likelihood): S-shape curve of x-marker, y-proportion of deaths ranging from 0-1. 
Derivatives should not be used to get Maximize Likelihood, because it might get a regional rather than the global Max. Instead, Iterative process should be used. 

## Assessment of the fitted model

**CI and tests (Wald test, LR test): **LR tests checks if the model increases the likelihood of obtaining the data compared to null model. 
**Goodness of fit of the model:** Chi-square test checks how far the observed p of outcome is away from the expected p of outcome predicted by logistic regression. 
**R square: ** Here, it is referred to as measures of effect size (how good exploratory variables are in predicting the outcome), not goodness of fit. 
**Discrimination**: How well a model distinguishes patients who survive from who die (AUC).Similar to diagnosis. 

**Validation:** How well the results will generalize to an independent data set. External validation is usually not feasible, but internal validation: **bootstrapping** (drawing samples with replacement, model may be developed in bootstrap samples and tested in the original sample), **cross-validation** (training set and testing set, performed over rounds; Brier score is used to estimate the prediction error). 

**Residual Analysis** Deviance Residual determines if individual points are not fitted by the model; Partial residuals

## Variable selection

**Forward selection** Initial model does not include any covariables. It includes covariables that decrease AIC.
**Backward selection** Initial model includes all covariables. It excludes covariables that decrease AIC.
**Stepwise selection** 

## polychotomous scale, continuous variable and interaction
One-degree FP function, Two-degree FP function
Quantitative interaction and Qualitative interaction

# Classification: Random forest and maschine learning
idea of underfitting, overfitting and trade off. 
## Random forest
many decision trees, and the final classification would be the mode of all results from trees. 
other ML methods: naive bayes, decision tree, linear SVM, nearest neighbors


#Exercise 3

## Research Question:
How does the use of hard drug effect asthma among US citizens? 
Confounder: Age, SCS - income, education, work status; smoking and cannabis


## Data Exploration
```{r}
library(readr)
df <- read_delim("../../data/NHANES1.csv", delim = ";")


# dimensions, variable names, classes, and distributions, missing values. 
dim(df)
head(df)
colnames(df)
df %>% map(class)
summary(df) 

# variables like smoking status has the most missing value, but they are not real missing value.
```


```{r}
df %>% 
  ggplot(aes(x = asthma_ever)) + 
  geom_bar() +
  theme_bw() 

df %>% 
  ggplot(aes(x = asthma_now)) + 
  geom_bar() +
  theme_bw() 

df %>% 
  ggplot(aes(x = harddrg_ever)) + 
  geom_bar() +
  theme_bw() 

df %>% 
  ggplot(aes(x = age)) + 
  geom_histogram() 
  theme_bw() 

df %>% 
  ggplot(aes(x = factor(increl))) +  
  geom_bar() 
  theme_bw() 
  
df %>% 
  ggplot(aes(x = factor(educ))) +  
  geom_bar() 
  theme_bw() 

df %>% 
  ggplot(aes(x = factor(jobstat_lwk)))+ 
  geom_bar() 
  theme_bw() 
  
df %>% 
  ggplot(aes(x = factor(smokstat)))+
  geom_bar() 
  theme_bw() 
  
df %>% 
  ggplot(aes(x = factor(cannab_ever)))+
  geom_bar() 
  theme_bw()

gtsummary::tbl_cross(df, row = asthma_ever, col = harddrg_ever, missing = "no")

df %>% select(asthma_ever, asthma_now) %>% plyr::count()

```

## Data prparation
```{r}

# as.factor for categorical variables


df$increl <- as.factor(df$increl)
df$educ <- as.factor(df$educ)
df$jobstat_lwk <- as.factor(df$jobstat_lwk)
df$smokstat <-as.factor(df$smokstat)
df$cannab_ever <-as.factor(df$cannab_ever)
df$male <- as.factor(df$male)

df$asthma_ever <- as.factor(df$asthma_ever)
df$asthma_now <- as.factor(df$asthma_now)
df$harddrg_ever <- as.factor(df$harddrg_ever)

class(df$asthma_ever)
class(df$asthma_now)


data <- df %>% mutate(asthma_now_adjusted = case_when(
  is.na(asthma_now) & asthma_ever == FALSE ~ 0,    # here we replace NA using FALSE, because people might only answer asthma_ever but not asthma_now. 
  asthma_now == FALSE ~ 0,
  asthma_now == TRUE ~ 1
),
phq_sum = phq1 + phq2 + phq3 + phq4 + phq5 + phq6 + phq7 +phq8 + phq9,
phq_average = (phq_sum - mean(phq_sum,na.rm = TRUE))/sd(phq_sum,na.rm = TRUE)
)
# remember to remove na when calculate the standardised value




```

## Modeling
```{r}
summary(data$asthma_now_adjusted)

m0 <- glm(asthma_now_adjusted ~ 1, data = data, family = binomial(link = "logit"))
dim(df)
nobs(m0) 
summary(m0)  
boot::inv.logit(m1$coefficients[1]) 


m1 <- glm(asthma_now_adjusted ~ harddrg_ever, data = data, family = binomial(link = "logit"))

nobs(m1) # 3481
summary(m1)
boot::inv.logit(m1$coefficients[2]) 
gtsummary::tbl_cross(data, row = asthma_now_adjusted, col = harddrg_ever, missing = "no")


m2 <- glm(asthma_now_adjusted ~ harddrg_ever + age + increl + educ + jobstat_lwk + smokstat + cannab_ever + phq_average, data = data, family = binomial)

m2 %>% tidy(conf.int = TRUE, exponentiate = TRUE)

```




















