---
title: "Day 6 Advanced poisson regression"
output: html_document
---

1.Start with poisson
2.Check overdispersion using plots
3.Use following solutions if there is oversdispersion


# Quasipoisson
```{r}

qm <- glm(Freq ~ Sat, data = MASS::housing, family = quasipoisson())

summary(qm)
summary(qm)$dispersion
# (Dispersion parameter for quasipoisson family taken to be 12.65788)
# The variance is 12.65788 times of mean. 

AIC(qm) 
# NA, cannot use AIC because the likelihood is not specified. 

```


# Negative-Binomial

A combination of poisson and gamma distribution.
Use gamma distribution to describe lambda, rather than using only one lambda.

```{r}

# MASS:: glm.nb()
nb <- glm.nb(Freq ~ Sat, data = MASS::housing, link = log)
AIC(nb)

```

# Zero-Inflated Poisson

Add a binary indicator, which defines if an observation's count is 0 or follows Poisson. This binary indicator is estimated by logistic regression. 
**Which count data variables with excess zero?**
Times of using hard drug
Number of birth

```{r}
# pscl::zeroinfl()
# install.packages("pscl")

zip <- pscl::zeroinfl(doctor ~ 1, data = Ecdat::Doctor, dist = "poisson" )
summary(zip)

broom.helpers::tidy_zeroinfl(zip)
```

# Zero-inflated Negative-Binomial Regression

```{r}

zip <- pscl::zeroinfl(doctor ~ 1, data = Ecdat::Doctor, dist = "negbin" )
summary(zip)

```

# Live coding

```{r}

library(conflicted)
library(tidyverse)
library(magrittr)
library(gtsummary)
library(ggplot2)
library(broom)
library(broom.helpers)
library(pscl)
library(Ecdat)
library(patchwork) # combine several plots

conflicts_prefer(
  dplyr::select(),
  dplyr::filter()
)
conflict_scout() # check the conflicts

d <- Ecdat::Doctor
d%>% map(class)

```
## Data Exploration
```{r}
d %>% 
  ggplot(aes(x = doctor)) + 
  geom_histogram(binwidth = 1) + 
  theme_bw()
# zero access: many people don't go to the docter within a 4 month time frame.

d %>% 
  tbl_summary() 
# no missing value

d %<>%
  mutate(
    access_std = (access - mean(access))/sd(access)
  )

mean(d$access)
sd(d$access)

```

## Modelling
### Poisson
```{r}
m0 <- glm(doctor ~ 1, family = poisson(), data = d)
summary(m0)

# poisson
m_po <- glm(doctor ~ health + access_std + children, data = d, family = poisson() )
summary(m_po)
nobs(m_po)

#Use residual diagnostics for detection of outliers and influential data points
plot(m_po)

sim_po <- simulate(m_po, nsim = 5)
cbind(d$doctor, sim_po) %>% View() #simulated data does not fit very well. 


ggplot() +
  geom_density(data = sim_po, aes(x = sim_1), color = "red")+
  geom_density(data = sim_po, aes(x = sim_2), color = "green")+
  geom_density(data = sim_po, aes(x = sim_3), color = "blue")+
  geom_density(data = sim_po, aes(x = sim_4), color = "orange")+
  geom_density(data = sim_po, aes(x = sim_5), color = "purple")+
  geom_density(data = d, aes(x = doctor), linewidth = 1)+
  theme_bw()


# If there is overdispersion, the variance of real data should be larger than the data simulated by model. 

plot_data_var <- data.frame(sim_var = sim_po %>% 
  map(var) %>% 
  unlist()) # A data frame should include a vector

plot_data_var %>% 
  ggplot() +
  geom_histogram(aes(x = sim_var)) +
  geom_vline(aes(xintercept = var(d$doctor) ), color = "red") +
  theme_bw()

plot_data_mean <- data.frame(sim_mean = sim_po %>% 
  map(mean) %>% 
  unlist()) # A data frame should include a vector

plot_data_mean %>% 
  ggplot() +
  geom_histogram(aes(x = sim_mean)) +
  geom_vline(aes(xintercept = mean(d$doctor) ), color = "green") +
  theme_bw()

# check access zero
sum(sim_po$sim_1 == 0)
sim_po %>%
  equals(0) %>%
  colSums()

sum(d$doctor== 0) # more zeros than the simulated data

plot_data_zeros <- data.frame(sim_zeros = 
                              sim_po %>%
                                equals(0) %>%
                                colSums())

plot_data_zeros %>%
  ggplot()+
  geom_histogram(aes(x = sim_zeros))+
  geom_vline(aes(xintercept = sum(d$doctor==0) ), color = "blue") +
  theme_bw()
  
# the plot tells us that the data from model contains more zeros than expected


```

### Quasipoisson
```{r}

m_qpo <- glm(doctor ~ health + access_std + children, data = d, family = quasipoisson())

```


### Negative Binomial
```{r}

m_nb <- glm.nb(doctor ~ health + access_std + children, data = d )
summary(m_nb)

jtools::plot_summs(m_po,m_nb, exp = TRUE)
# the confidence interval of nb model is larger than the poisson model. 
# The poisson estimates the variance too small, so corrected in nb model. 

```

### Zero-inflated
```{r}

m_zip <- zeroinfl(doctor ~ health + access_std + children, data = d, dist = "poisson")
summary(m_zip)

```

### Zero-inflated negative-binomial regression
```{r}
# pscl::zeroinfl

m_zipb <- zeroinfl(doctor ~ health + access_std + children, data = d, dist = "negbin")
summary(m_zipb)

```

### Compare models
```{r}

AIC(m0, m_po,m_qpo, m_nb, m_zip, m_zipb)
# not working by anova
```


# Exercise 6
## Modeling
### Poisson
```{r}
d1 <- MASS::quine

m_po1 <- glm(Days ~ Eth + Sex + Age + Lrn, data = d1, family = poisson())
tbl_regression(m_po1, exponentiate = TRUE)


residual <- residuals(m_po1)
plot(m_po1)
d[104, ]
d[72, ]
d[59, ]


m_po1  %>% 
  simulate(nsim = 100) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  ) %>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") + # color = sim, multiple colors
  geom_density(
    data = d1,
    mapping = aes(x = Days),
    linewidth = 1
  ) +
  theme_bw()
# the fit is not very good so far

sim_po1 <- simulate(m_po1, nsim = 100)
# cbind(d$Days, sim_po1$sim_7) %>% View() 


plot_data_var1 <- data.frame(sim_var1 = sim_po1 %>% 
  map(var) %>% 
  unlist()) # A data frame should include a vector

plot_data_var1 %>% 
  ggplot() +
  geom_histogram(aes(x = sim_var1)) +
  geom_vline(aes(xintercept = var(d1$Days) ), color = "red") +
  theme_bw()


```



### Quasipoisson
```{r}
# Fit a quasipoisson model and extract the estimated dispersion parameter. What does the value mean?


m_qpo1 <- glm(Days ~ Eth + Sex + Age + Lrn, data = d1, family = quasipoisson())
summary(m_qpo1)
summary(m_qpo1)$dispersion

# (Dispersion parameter for quasipoisson family taken to be 13.16691)
# 13.16691 - The variance is 13.16691 times of mean? 


```

### Negative Binomial

```{r}

m_nb1 <- glm.nb(Days ~ Eth + Sex + Age + Lrn, data = d1) # we don not assign value for family
summary(m_nb1)
m_nb1 %>% tbl_regression(exponentiate = TRUE)

#Create 100 simulations of the distribution of number of absent school days based on the negative-binomial model and plot them together with the observed data. Compare the plot to the one from exercise 6.1.3.


m_nb1  %>% 
  simulate(nsim = 100) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  ) %>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") + # color = sim, multiple colors
  geom_density(
    data = d1,
    mapping = aes(x = Days),
    linewidth = 1
  ) +
  theme_bw()


```


```{r}

sim_po2 <- simulate(m_nb, nsim = 100)
plot_data_var2 <- data.frame(sim_var2 = sim_po2 %>% 
  map(var) %>%  # map applys to every variable. 
  unlist())     # A data frame should include a vector.

plot_data_var2 %>% 
  ggplot() +
  geom_histogram(aes(x = sim_var2)) +
  geom_vline(aes(xintercept = var(d$Days) ), color = "red") +
  theme_bw()


```

Compare models
```{r}
AIC(m_po1, m_nb1)

# negative binomial distribution has the better fit. 

```


### Zero-inflated negative-binomial regression
```{r}

m_zipb1 <- zeroinfl(Days ~ Eth + Sex + Age + Lrn, data = d1, dist = "negbin")
summary(m_zipb1)
broom.helpers::tidy_zeroinfl(m_zipb1, exponentiate = TRUE) 
# same result, we do not use tbl_regression, not suitable for this. 


AIC(m_po1, m_nb1,m_zipb1)


# This Zero-inflated negative-binomial regression does not improve the fit.
```

### Zero-inflated binomial
```{r}
m_zipp1 <- zeroinfl(Days ~ Eth + Sex + Age + Lrn, data = d1, dist = "poisson")

AIC(m_po1, m_nb1,m_zipb1,m_zipp1 )

# not better than m_zipb1
```

Research Question:
1. Does ethnic background, sex, age and Learner status affect the days absent from schools in the year?

Interpretation: Compared to female, male have 18% more days of absence from school. This effect is not significant.


# Freestyle: OPF “Demand for Medical Care by the Elderly."

**Outcome group: **
ofp, ofnp, opp, opnp, emr, hosp ("number of ... ", count data variable)

**Predictor group: **
numchron, adldiff, age, black, sex, maried, school, faminc, employed, privins, medicaid, region, hlth

**Research Question**
How does all predictors effect the number of hospitalizations (aka. demand for Medical Care)?


## Data Exploration
```{r}

d2 <- Ecdat::OFP

view(d2)
nrow(d2)
ncol(d2)
dim(d2)

d2 %>% tbl_summary()

d2 %>%
  is.na() %>%
  colSums() %>%
  sort(decreasing = TRUE)

d2 %>% map(class)

# perfect data, no missing value, perfect data type
# no Data Preparation needed for this case
```

## Modeling
### Poisson

```{r}


m_po2 <- glm(hosp ~ numchron + adldiff + age + black + sex + maried + school + faminc + employed + privins + medicaid + region + hlth, data = d2, family = poisson())

tbl_regression(m_po2, exponentiate = TRUE)

# numchron: number of chronic conditions
# adldiff: the person has a condition that limits activities of daily living ?
# school: number of years of education
# privins: is the person covered by private health insurance?
# medicaid : is the person covered by medicaid ?
# hlth: self-perceived health (excellent, poor, other)

plot(m_po2)

m_po2  %>% 
  simulate(nsim = 100) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  ) %>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") + # color = sim, multiple colors
  geom_density(
    data = d2,
    mapping = aes(x = hosp),
    linewidth = 1
  ) +
  theme_bw()
# the poisson fits relativly well, but let's see if there is better option

sim_po2 <- simulate(m_po2, nsim = 100)
# cbind(d$Days, sim_po1$sim_7) %>% View() 


plot_data_var2 <- data.frame(sim_var2 = sim_po2 %>% 
  map(var) %>% 
  unlist()) 

plot_data_var2 %>% 
  ggplot() +
  geom_histogram(aes(x = sim_var2)) +
  geom_vline(aes(xintercept = var(d2$hosp) ), color = "red") +
  theme_bw()

# the variance simulated data is far away from the real data


plot_data_zeros2 <- data.frame(sim_zeros2 = 
                               sim_po2 %>%
                               equals(0) %>%
                               colSums())

plot_data_zeros2 %>%
  ggplot()+
  geom_histogram(aes(x = sim_zeros2))+
  geom_vline(aes(xintercept = sum(d2$hosp==0) ), color = "blue") +
  theme_bw()

# more zeros in reality than in the simulated data. 

```

### Quasipoisson
```{r}

m_qpo2 <- glm(hosp ~ numchron + adldiff + age + black + sex + maried + school + faminc + employed + privins + medicaid + region + hlth, data = d2, family = quasipoisson())

summary(m_qpo2)
summary(m_qpo2)$dispersion


# (Dispersion parameter for quasipoisson family taken to be 1.589551)
# variance is only 1.589 times of the mean. They are not very different from each other
# No need to correct the modal by making it more complex.
```


### Negative Binomial
```{r}

m_nb2 <- glm.nb(hosp ~ numchron + adldiff + age + black + sex + maried + school + faminc + employed + privins + medicaid + region + hlth, data = d2)

AIC(m_po2, m_nb2)


m_nb2  %>% 
  simulate(nsim = 100) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "sim",
    values_to = "val"
  ) %>% 
  ggplot() +
  geom_density(aes(x = val, group = sim), color = "grey") + # color = sim, multiple colors
  geom_density(
    data = d2,
    mapping = aes(x = hosp),
    linewidth = 1
  ) +
  theme_bw()

# the poisson fits relatively well, but let's see if there is better option

sim_nb2 <- simulate(m_nb2, nsim = 100)
# cbind(d$Days, sim_po1$sim_7) %>% View() 


plot_data_var2 <- data.frame(sim_var2 = sim_nb2 %>% 
  map(var) %>% 
  unlist()) 

plot_data_var2 %>% 
  ggplot() +
  geom_histogram(aes(x = sim_var2)) +
  geom_vline(aes(xintercept = var(d2$hosp) ), color = "red") +
  theme_bw()

jtools::plot_summs(m_po2,m_nb2, exp = TRUE)
```

### Zero-inflated negative-binomial regression

```{r}

m_zipb2 <- zeroinfl(hosp ~ numchron + adldiff + age + black + sex + maried + school + faminc + employed + privins + medicaid + region + hlth + , data = d2, dist = "negbin")

AIC(m_po2, m_nb2, m_zipb2)




# This last model is the best! 


```

```{r}

summary(m_zipb2)
broom.helpers::tidy_zeroinfl(m_zipb2, exponentiate = TRUE) 

```


### Zero-inflated poisson
```{r}

m_zipp2 <- zeroinfl(hosp ~ numchron + adldiff + age + black + sex + maried + school + faminc + employed + privins + medicaid + region + hlth, data = d2, dist = "poisson")

AIC(m_po2, m_nb2, m_zipb2, m_zipp2)

# This is not better than m_zipb2

```

