---
title: "Day 8"
output: html_document
---


**Hazard** is the probability of death at time t -- this is only a density function, different from Hazard function. "Most of the people died at the age around 90."

**Hazard function** measures the probability of death at time t, given that the individual survived up to time t. / The hazard is the risk of an event at time t, given the person is event-free before t. "But the risk of dying is increasing after 90."
H(t) = - ln(S(t))

In the **exponential survival** model, the hazard is assumed to be the constant over time. 
S(t) = exp(-h*t)

**Proportional Hazard model** is used for comparing groups/controlling for covariates. 
Assumptions: 
The Hazards Ratio remains constant, independent of time.
The effects of covariates are constant over time. 
(If the assumptions are violated, the cox model might not be the best.)

Interpretation: 
HR = 2.61: The risk of death is increased by a factor of 2.61. 

Model Comparison:
Can us AIC or partial F test. 

# Aderson data
```{r}

library(survival)
library(survminer)
library(ggplot2)
library(MASS)
library(gtsummary)
library(dplyr)
library(conflicted)
library(readr)

conflicts_prefer(
  dplyr::select(),
  dplyr::filter()
)
conflict_scout() # check the conflicts


```

## Data Exploration and Preparation
```{r}

d <- autoReg::anderson 
?autoReg::anderson 
d %<>% 
  mutate(sex = case_when(
  sex == "0" ~ "female",
  sex == "1" ~ "male"
),
sex = factor(sex,
  levels = c("female", "male")
))

d %<>% 
  mutate(rx = case_when(
  rx == "1" ~ "control",
  rx == "0" ~ "treatment"
),
  rx = factor(rx, 
    levels = c("control","treatment")
    ))

class(d$sex)
m_cox <- coxph(Surv(time,status) ~ sex + rx + logWBC, data = d)
m_cox$n
m_cox$nevent
summary(m_cox) # n= 42, number of events= 30 
m_cox %>% tbl_regression(exponentiate = TRUE)


```

## Survival Analysis 
```{r}


# Check if there is a linear relation between log hazard and continuous predictors.
# only one continuous predictor here logWBC

# method 1
ggcoxfunctional(Surv(time,status) ~ logWBC, data = d, ggtheme = theme_bw())

# method 2
data.frame(
  resid = residuals(coxph(Surv(time, status) ~ 1, data = d),
  type = "martingale"
  ),
  val = d$logWBC
)%>%
  ggplot(aes(x = val, y = resid)) +
  geom_point() +
  stat_smooth(method = "loess") +
  theme_bw() +
  ylim(-2, 2) +
  ylab("Martingale Residuals \nof Null Cox Model") +
  xlab("logWBC")


# Evaluate if the proportional hazard assumption holds.
ggcoxzph(cox.zph(m_cox))


```


## Visualisation
```{r}

# Plot the modeled survival curves together with Kaplan-Meier curves. 
# Plot two curves each, representing treated and controls. 
# Use the mean value of logWBC for the modelled curves. Create separate plots for men and women. 
# Use argument subset = (sex == "male")` in survfit() to get sex-specific KM curves.

# Create separate newdata for men and women
newdata_male <- data.frame(
  rx = c("control", "treatment"),
  sex = rep("male", 2),
  logWBC = rep(mean(d$logWBC), 2)
)
rownames(newdata_male) <- c("rx=0", "rx=1")

newdata_female <- data.frame(
  rx = c("control", "treatment"),
  sex = rep("female", 2),
  logWBC = rep(mean(d$logWBC), 2)
)
rownames(newdata_female) <- c("rx=0", "rx=1")

# Create separate curves for men and women
modelled_male <- survfit(m_cox, newdata = newdata_male, data = d)
km_male <- survfit(Surv(time, status) ~ rx, data = d, subset = (sex == "male"))

modelled_female <- survfit(m_cox, newdata = newdata_female, data = d)
km_female <- survfit(Surv(time, status) ~ rx, data = d, subset = (sex == "female"))

# Create separate plots for men and women
plot_men <- ggsurvplot_combine(list(CoxPH = modelled_male, KM = km_male),
  data = d, ggtheme = theme_bw()
)
plot_men$plot +
  labs(title = "Men")

plot_women <- ggsurvplot_combine(list(CoxPH = modelled_female, KM = km_female),
  data = d, ggtheme = theme_bw()
)
plot_women$plot +
  labs(title = "Women")


```

# Free style: nepal data

Research Question: Do infants of teenage mother has a higher risk of death?

## Data Exploration

```{r}
d2 <- read_delim("../../data/nepalDHS.csv", delim = ",") 
d2 %>% tbl_summary()
view(d2)

```


## Data Preparation
```{r}
# I recoded all variables just to keep myself focused. 
d2 %<>% 
  mutate(male = case_when(
  Male == "1" ~ "male",
  Male == "0" ~ "female"
),
male = factor(male,
  levels = c("female", "male")
))


d2 %<>% 
  mutate(urban = case_when(
  Urban == "1" ~ "urban",
  Urban == "0" ~ "rural"
),
 urban = factor(urban,
  levels = c("urban", "rural")
))


d2 %<>% mutate(wealth = case_when(
  Wealth == 1 ~ "poorest",
  Wealth == 2 ~ "poor",
  Wealth == 3 ~ "medium",
  Wealth == 4 ~ "rich",
  Wealth == 5 ~ "richest",
), 
wealth = factor(wealth, 
                   levels = c("poorest","poor","medium", "rich","richest"))
)


d2 %<>% 
  mutate(first = case_when(
  First == "1" ~ "first born infant",
  First == "0" ~ "not first born infant"
),
 first = factor(first,
  levels = c("first born infant", "not first born infant")
))


d2 %<>% 
  mutate(alive = case_when(
  Alive == "1" ~ "alive",
  Alive == "0" ~ "dead"
),
  alive = factor(alive, 
    levels = c("alive","dead")
    ))

d2 %<>% 
  mutate(first_dead = case_when(
  FirstInfantDeath == "1" ~ "first infant death in family",
  FirstInfantDeath == "0" ~ "subsequent infant death in family"
),
  first_dead = factor(first_dead, 
    levels = c("first infant death in family","subsequent infant death in family")
    ))


d2 %<>%
  mutate(motherage = case_when(
    MotherAge < 20 ~ "<20y",
    MotherAge >= 20 ~ ">=20y"
  ),
  motherage = factor(motherage,
                     levels = c("<20y",">=20y")))


view(d2)


```

## Survival Analysis

```{r}


```

