---
title: "Day 1 Workflow"
output: html_document
---

Workflow:

**Research Question**
**They guide your analysis! 
**PICO: P-population, I-intervention, C-comparison with another intervention, O - outcome of interests, T - time frame

**Data Exploration** 
** Purpose? structure of data, errors of data , problems (e.g., complexities) of data, define tasks for data preparation, get a feeling about the data. 
** Tasks? Frequencies of variables, summaries, distributions (e.g, histogram), number of missing values, variable types, correlations. 
** Tools? Tables, Plots, Values.
** Tips? look at all relevant variables, get number of NAs (check the pattern, but take no action for now), summary, plot (density, histogram), repeat within important subgroups. 

**Data Preparation**  Get a clean set before running your analysis! 
** Tasks: Combine, recode (e.g., from character to numeric), and transform variables (log10 to make variable more normally distributed, standardization)
** Tips? check if it works when every you add a new variable. 

**Modeling** Do not forget to check the literature
** Initial: start with the simplest model (e.g., poisson model)
** Fitting: e.g, calculation R
** Evaluate: understand your variables (e.g., IRR-Incidence Rate Ratio, 95%CI, p-value), prediction, diagnostics (if assumptions are violated)
** Compare: model comparison using e.g., AIC ()
** Modify: extend your model by adding complexity as needed, based on diagnostics and domain knowledge. (e.g., switch from poisson to negative-binomial)

**Interpretation**
Answer your questions, limitations, strengths, bias, random error, generalisability, comparison with existing literature, open questions. 

**Visualization** Across all steps in the workflow! could visulaize for your self to better understand or for others. 


```{r}
install.packages(
  c("autoReg",
    "boot",
    "broom",
    "broom.helpers",
    "conflicted",
    "Ecdat",
    "gglm",
    "gtsummary",
    "jtools",
    "magrittr",
    "mgcv",
    "MASS",
    "patchwork",
    "plyr",
    "pscl",
    "survival",
    "survminer",
    "tidyverse",
    "visreg"),
  dependencies = TRUE
)


```

Creating vectors and data.frames
```{r}

seq(3,-1)
seq(8)
seq(from = 1, to =5, by = 0.5)
seq(from = 1, to =5, length.out = 5)
rep(1, times = 4)
rep(1, 4)
rep(1:3, times = 4)
data.frame(var1 = 1:3, var = c(TRUE, FALSE, TRUE), var3= rep(NA, 3))

```


Exercise: Creating vectors and data.frames
```{r}
# 1.1.1
seq(from = 0, to = 1, length.out = 100)

# 1.1.2
seq(from = 20, to = 2000, by = 20)

# 1.1.3
rep(c("A", "B", "C","D"), each = 25)

# 1.1.4
rep(c(TRUE, FALSE), each = 50) 

# 1.1.5
data.frame(var1 = seq(from = 0, to = 1, length.out = 100), 
           var2 = seq(from = 20, to = 2000, by = 20), 
           var3 = rep(c("A", "B", "C","D"), each = 25),
           var4 = rep(c(TRUE, FALSE), each = 50))

# 1.1.6
data.frame(var1 = seq(from = 0, to = 10), 
           var2 = rep(c("a", "b"), each = 5),
           var4 = seq(from = 0, to = 1 , by = 0.1))
# Answer: there is error because there are different arguments in different rows.

data.frame(var1 = seq(from = 0, to = 9), 
           var2 = rep(c("a", "b"), each = 5),
           var4 = seq(from = 0, to = 0.9 , by = 0.1))


```

Exercise: Factors
```{r}

# 1.2.1
library(ggplot2)

manufacturer_factor = as.factor(mpg$manufacturer)
summary(manufacturer_factor)
levels(manufacturer_factor) 
# if want to control : factor (manufacturer_factor, c( ....))
# can also change for levels and labels, change order, specifically for creating plot


library(forcats)
levels(as.factor(mpg$drv))
drv_factor <- factor(mpg$drv,
                     levels = c("f", "r", "4"),
                     labels = c("front-wheel drive","rear wheel drive","4wd"))

fct_rev(drv_factor)
summary(drv_factor)

v <- fct_infreq(drv_factor) 
summary(v)

```

Exercise: Piping

```{r}
library(dplyr)

#1.3.1
x1 <- seq(from = 10, to =30, length.out = 10)
x2 <- sum(x1) # solution without piping 
x2
x3 <- seq(from = 10, to =30, length.out = 10) %>% sum()
x3 # solution with piping 

#1.3.2
tot_na <- is.na(txhousing)  # tot_na <- is.na(txhousing$sales) also works
colsums_tot_na <- colSums(tot_na)
colsums_tot_na_sorted <- sort(colsums_tot_na)
colsums_tot_na_sorted["sales"] 

sort(colSums(is.na(txhousing))) # without piping
txhousing %>% is.na() %>% colSums() %>% sort() #with piping

#1.3.3
colsums_tot_na_sorted["city"] 
colsums_tot_na_sorted["volume"] 

#1.3.4 

sum(txhousing$sales,na.rm =TRUE)
txhousing$sales %>% sum(na.rm = TRUE)


```

Exercise: dplyr

```{r}
# Extract rows:filter()
#1.4.1

df <- diamonds %>% filter(cut == "Good") 
df

#1.4.2
df <- df %>% select(price, x,y,z) %>% mutate(price_cat = case_when (
  price < 1000 ~ "cheap",
  price >= 10000 ~ "expensive", 
  price >= 1000 & price < 10000 ~ "normal"     # or .default = ~ "normal"
),
price_cat = as.factor(price_cat)) 

df$price_cat

# 1.4.3

df <- df %>% mutate(price_euro = price*0.86) %>% dplyr::filter(price_euro > 10000) 
df

# 1.4.4
df %>% mutate(s = x+y+z)
# or
diamonds %>%  rowwise () %>% mutate(s = sum (x,y,z))


```

Exercise: ggplot
```{r}

library(ggplot2)

ggplot(data = diamonds %>% filter(cut == "Good")) +
  geom_point(aes(x = carat, y = price, color=clarity),shape = "square") + theme_bw()+ 
  labs( title = "Plot title",
        subtitle = "A subtitle",
        caption = "Small text below the plot",
        tag = "Short text in top-left corner",
        x = "New x-axis title",
        y = "New y-axis title",
        color = "Title of color legend")+
  facet_wrap(~ color)

ggplot(data = diamonds) + geom_boxplot(aes(x = cut, y = price, color = cut))+
      labs(subtitle = "A subtitle",
        caption = "Small text below the plot",
        x = "Price in $",
        y = "Quality of the cut",
        color = "Price in $")


```

Exercise: Descriptive

```{r}

install.packages("plyr")

#1.6.1
mpg %>% 
  select(cty, hwy,fl) %>% 
  gtsummary::tbl_summary() # if pass to a df, df is a table rather than a dataframe

#1.6.2
mpg %>% select(cty, hwy,fl, drv) %>% 
  gtsummary::tbl_summary(include = c(cty, hwy,fl), by =drv) %>%
  gtsummary::add_overall()

#1.6.3
mpg$drv %>% plyr::count()

#1.6.4
mpg %>% 
  select(drv, fl) %>% 
  plyr::count()

```

